---
title: "Comparative analysis of asexual structures morphology and culture growth in _Epichloë_"
author: "Sotomayor-Alge, Alba"
date: "02-02-2026"
output:
  bookdown::pdf_book:
    toc: yes
    toc_depth: 3
    number_sections: yes
    highlight: tango
    latex_engine: xelatex
    keep_tex: true
---

\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Working session setup

```{r Load packages into working session}

# Function to silently load packages without warnings or messages
load_packages <- function(pkgs) {
  invisible(
    lapply(pkgs, function(pkg) {
      suppressWarnings(
        suppressMessages(
          if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
            stop(paste("❌ Package not found:", pkg), call. = FALSE)
          }
        )
      )
    })
  )
}

# Vector of required packages
packages_needed <- c("dplyr", "tidyr", "ggplot2", "knitr", "ggcorrplot", "vegan", 
                     "coin", "FactoMineR", "factoextra", "patchwork", "MASS", 
                     "caret", "lme4", "lmerTest", "emmeans", "openxlsx", "rlang",
                     "performance", "stringr", "multcompView", "multcomp"
)

# Call the function to load them
load_packages(packages_needed)

```

```{r Working session information}

# Print the information of the working session in order to ensure reproducibility
sessionInfo()

```

\newpage

```{r Create directories and exporting function}

# Create directories within the file system to save the results of the analyses
dir.create("G:/ESPORAS_SPCOMP/outputs_script1", recursive = TRUE, showWarnings = FALSE)
dir.create("G:/ESPORAS_SPCOMP/outputs_script1/1_Spores_dataset", recursive = TRUE, 
           showWarnings = FALSE)
dir.create("G:/ESPORAS_SPCOMP/outputs_script1/1_Spores_dataset/1_EDA", recursive = TRUE, 
           showWarnings = FALSE)
dir.create(".G:/ESPORAS_SPCOMP/outputs_script1/1_Spores_dataset/2_STAT", recursive = TRUE, 
           showWarnings = FALSE)
dir.create("G:/ESPORAS_SPCOMP/outputs_script1/1_Spores_dataset/3_PCA", recursive = TRUE, 
           showWarnings = FALSE)
dir.create("G:/ESPORAS_SPCOMP/outputs_script1/1_Spores_dataset/4_LDA", recursive = TRUE, 
           showWarnings = FALSE)
dir.create("G:/ESPORAS_SPCOMP/outputs_script1/2_Culture_growth_dataset", recursive = TRUE, 
           showWarnings = FALSE)

# Helper function to add and format sheets of excel files created in this script
add_formatted_sheet <- function(wb, sheet_name, data) {
  addWorksheet(wb, sheet_name)
  writeData(wb, sheet = sheet_name, x = data)

  # Define styles
  style_text <- createStyle(
    border = "TopBottomLeftRight",
    halign = "center",
    valign = "center"
  )
  style_integer <- createStyle(
    border = "TopBottomLeftRight",
    halign = "center",
    valign = "center",
    numFmt = "0"
  )
  style_decimal <- createStyle(
    border = "TopBottomLeftRight",
    halign = "center",
    valign = "center",
    numFmt = "#,##0.000"
  )

  # Apply style to each column individually
  for (colname in names(data)) {
    col_index <- which(names(data) == colname)
    column_data <- data[[colname]]

    if (is.numeric(column_data)) {
      if (is.integer(column_data)) {
        # True integer
        addStyle(wb, sheet = sheet_name, style = style_integer,
                 rows = 1:(nrow(data) + 1), cols = col_index, gridExpand = TRUE)
      } else {
        # Numeric (with decimals)
        addStyle(wb, sheet = sheet_name, style = style_decimal,
                 rows = 1:(nrow(data) + 1), cols = col_index, gridExpand = TRUE)
      }
    } else {
      # Non-numeric columns
      addStyle(wb, sheet = sheet_name, style = style_text,
               rows = 1:(nrow(data) + 1), cols = col_index, gridExpand = TRUE)
    }
  }

  # Adjust column widths
  setColWidths(wb, sheet = sheet_name, cols = 1:ncol(data), widths = "auto")
}

```

\newpage

# Morphological analisys of the asexual reproductive structures (conidia and conidiophores) of _Epichloë_ endophytes

## Load, explore & mutate data set

```{r Data set load, exploration and mutation}

# Load data set considering its inner structure
spores_data <- read.table("G:/ESPORAS_SPCOMP/MORPHOLOGY/spores_dataset_spcomp_final.csv",
                          dec = ",", sep = ";", header = TRUE)

# Sanity check. Ensures that all required columns are present before the analysis 
# continues. If any required column is missing, execution stops with an error message.
required_cols <- c("Species","SampleID","Ind","Rep","conidL","conidW","conidiophL",
                   "conidiophW","conidA")
missing <- setdiff(required_cols, colnames(spores_data))
if (length(missing) > 0) stop("Missing columns: ", paste(missing, collapse=", "))

# Convert categorical variables to factors and create PlateID from Ind × Rep
spores_data <- spores_data %>%
               mutate(Species = as.factor(Species), SampleID = as.factor(SampleID),
                      Ind = as.factor(Ind), Rep = as.factor(Rep),
                      PlateID = interaction(Ind, Rep, drop = TRUE))

# Create data set containing excluding the outgroup species (Farun)
spores_data_ingroup <- spores_data %>% filter(Species != "Farun")

# Add conidial L/W ratio (shape metric)
spores_data <- spores_data %>%
               mutate(conidLW = ifelse(is.na(conidW) | conidW == 0, NA_real_, 
                                       conidL / conidW))

spores_data_ingroup <- spores_data_ingroup %>%
                       mutate(conidLW = ifelse(is.na(conidW) | conidW == 0, NA_real_, 
                                               conidL / conidW))

# Define morphological variables used in analyses and select variables for PCA
morpho_vars_all <- c("conidL","conidW","conidiophL","conidiophW","conidA", "conidLW")
morpho_vars_stats <- c("conidL","conidW","conidiophL","conidiophW","conidA")
morpho_vars_permpca <- c("conidL","conidW","conidiophL","conidiophW")

# Create aggregated data sets: (1) plate-level means (biological/technical replicate)
# and (2) isolate-level means (biological isolate)
plate_means <- spores_data %>%
               group_by(Species, Ind, Rep) %>%
               summarise(across(all_of(morpho_vars_all), ~mean(.x, na.rm = TRUE)),
               .groups = "drop")
plate_means_ingroup <- spores_data_ingroup %>%
               group_by(Species, Ind, Rep) %>%
               summarise(across(all_of(morpho_vars_all), ~mean(.x, na.rm = TRUE)),
               .groups = "drop")
ind_means <- plate_means %>%
             group_by(Species, Ind) %>%
             summarise(across(all_of(morpho_vars_all), ~mean(.x, na.rm = TRUE)),
             .groups = "drop")
ind_means_ingroup <- plate_means_ingroup %>%
             group_by(Species, Ind) %>%
             summarise(across(all_of(morpho_vars_all), ~mean(.x, na.rm = TRUE)),
             .groups = "drop")

```

Before continuing with the analyses, the existence of outliers is assessed applying the IQR method.

```{r Outliers removal (IQR method)}

# Identify IQR-based outliers per morphological variable (flag only, no deletion)
# considering the data set that includes only E. festucae.
flag_outliers_iqr <- function(x) {
  q1 <- quantile(x, 0.25, na.rm = TRUE)
  q3 <- quantile(x, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  lower <- q1 - 1.5*iqr
  upper <- q3 + 1.5*iqr
  (x < lower) | (x > upper)
}

spores_data_flagged <- spores_data_ingroup
for (v in morpho_vars_all) {
  spores_data_flagged[[paste0(v, "_out")]] <- flag_outliers_iqr(spores_data_flagged[[v]])
}

# Summary table with the number of outliers per morphological variable
outlier_summary <- sapply(paste0(morpho_vars_all, "_out"),
                          function(v) sum(spores_data_flagged[[v]], na.rm = TRUE))

# As the number of outliers is quite low, all observations were kept for the following statistical
# analysis.
print(outlier_summary)

```
Given the low number of detected outliers, all observations were retained for the subsequent statistical analyses.

\newpage

## Exploratory Data Analysis (EDA)

### Graphic approach

```{r Visualize data distribution}

# Set output directory for exploratory boxplots
out_dir_eda <- "G:/ESPORAS_SPCOMP/outputs_script1/1_Spores_dataset/1_EDA/"
dir.create(out_dir_eda, recursive = TRUE, showWarnings = FALSE)

# Define colors for the study
species_col_ingroup <- c(Ftrich = "#e6ab02", Flamb = "#d95f02", Fyves = "#e72945",
                         Froth = "#283F79", Fnigr = "#1b9d99", Fprui = "#66a61e")
species_col <- c(Ftrich = "#e6ab02", Flamb = "#d95f02", Fyves = "#e72945",
                 Froth = "#283F79", Fnigr = "#1b9d99", Fprui = "#66a61e",
                 Farun = "#666666")

# Fix species order within the boxplots
spores_data_ingroup$Species <- factor(spores_data_ingroup$Species,
                                      levels = names(species_col_ingroup))
spores_data$Species <- factor(spores_data$Species,
                              levels = names(species_col))

# Create boxplots (considering only E. festucae)
for (v in morpho_vars_all) {
  p <- ggplot(spores_data_ingroup,
              aes(x = Species, y = .data[[v]], fill = Species)) +
       geom_boxplot(outlier.shape = NA,
                    width = 0.6,
                    alpha = 0.7,
                    linewidth = 0.3,
                    coef = 0,
                    col = "black") +
       stat_boxplot(geom = "errorbar",
                    width = 0.15,
                    linewidth = 0.25,
                    linetype = "dashed",
                    col = "black") +
       scale_fill_manual(values = species_col_ingroup) +
       theme_bw() +
       theme(panel.grid = element_blank()) +
       theme(legend.position = "none", axis.title.x = element_blank()) +
       ylab(v)
  ggsave(filename = file.path(out_dir_eda, paste0("boxplot_Ingroup_", v, ".pdf")),
         plot = p, width = 5, height = 4, units = "in")
}


# Create boxplots (considering all species)
for (v in morpho_vars_all) {
  p_allsp <- ggplot(spores_data,
          aes(x = Species, y = .data[[v]], fill = Species)) +
          geom_boxplot(outlier.shape = NA,
                       width = 0.6,
                       alpha = 0.7,
                       linewidth = 0.3,
                       coef = 0,
                       col = "black") +
          stat_boxplot(geom = "errorbar",
                       width = 0.15,
                       linewidth = 0.25,
                       linetype = "dashed",
                       col = "black") +
          scale_fill_manual(values = species_col) +
          theme_bw() +
          theme(panel.grid = element_blank()) +
          theme(legend.position = "none", axis.title.x = element_blank()) +
          ylab(v)
  ggsave(filename = file.path(out_dir_eda, paste0("boxplot_AllSpecies_", v, ".pdf")),
         plot = p_allsp, width = 5, height = 4, units = "in")
}

```

\newpage

### Basic statistics

```{r Basic statistics}

# Function for calculating basic statistics
BasicStatsGroupedLong <- function(data, group_vars, numeric_vars) {
  stats_long <- lapply(numeric_vars, function(var) {
    data %>%
      group_by(across(all_of(group_vars))) %>%
      summarise(n = as.integer(sum(!is.na(.data[[var]]))),
                Mean = mean(.data[[var]], na.rm = TRUE),
                Median = median(.data[[var]], na.rm = TRUE),
                SD = sd(.data[[var]], na.rm = TRUE),
                CV = ifelse(mean(.data[[var]], na.rm = TRUE) == 0, NA_real_,
                            sd(.data[[var]], na.rm = TRUE) / mean(.data[[var]], 
                                                                  na.rm = TRUE) * 100),
                Q1 = quantile(.data[[var]], 0.25, na.rm = TRUE),
                Q3 = quantile(.data[[var]], 0.75, na.rm = TRUE),
                IQR = IQR(.data[[var]], na.rm = TRUE),
                Min = min(.data[[var]], na.rm = TRUE),
                Max = max(.data[[var]], na.rm = TRUE),
                .groups = "drop") %>%
      mutate(Variable = var)}) %>%
      bind_rows() %>%
      relocate(Variable)

  return(stats_long)
}

# Statistics at species level
species_stats <- BasicStatsGroupedLong(spores_data, group_vars = "Species", 
                                       numeric_vars = morpho_vars_stats) %>%
  mutate(across(where(is.double), ~ round(.x, 3)))

# Statistics at individual/isolate level
ind_stats <- BasicStatsGroupedLong(spores_data, group_vars = c("Species", "Ind"), 
                                   numeric_vars = morpho_vars_stats) %>%
  mutate(across(where(is.double), ~ round(.x, 3)))

# Statistics at replicate level
rep_stats <- BasicStatsGroupedLong(spores_data, group_vars = c("Species", "Ind", "Rep"), 
                                   numeric_vars = morpho_vars_stats) %>%
  mutate(across(where(is.double), ~ round(.x, 3)))

# Export results to an excel file
wb_stats <- createWorkbook()
add_formatted_sheet(wb_stats, "Species_stats", species_stats)
add_formatted_sheet(wb_stats, "Ind_stats", ind_stats)
add_formatted_sheet(wb_stats, "Rep_stats", rep_stats)
saveWorkbook(wb_stats, file = file.path(out_dir_eda, "basic_stats_Allspecies.xlsx"),
             overwrite = TRUE)

```

\newpage

### Data distribution 

```{r Check normality of variables}

# Function for conducting normality tests
ShapiroTestGrouped <- function(data, group_vars, test_vars) {
    data %>%
    pivot_longer(cols = all_of(test_vars), names_to = "Variable", 
                 values_to = "Value") %>%
    group_by(across(all_of(group_vars)), Variable) %>%
    summarise(n = sum(!is.na(Value)),
              statistic = tryCatch(shapiro.test(Value)$statistic,
                                   error = function(e) NA_real_),
              p_value = tryCatch(shapiro.test(Value)$p.value, 
                                 error = function(e) NA_real_),
              .groups = "drop") %>%
    mutate(significance = ifelse(p_value > 0.05, "", "***"))
}

# Test normality considering all aggregation levels
shapiro_species <- ShapiroTestGrouped(spores_data, "Species", 
                                      morpho_vars_stats)
shapiro_ind <- ShapiroTestGrouped(spores_data, 
                                  group_vars = c("Species", "Ind"), 
                                  morpho_vars_stats)
shapiro_rep <- ShapiroTestGrouped(spores_data, 
                                  group_vars = c("Species", "Ind", "Rep"),
                                  morpho_vars_stats)

# Create an excel file
wb_norm <- createWorkbook()
add_formatted_sheet(wb_norm, "Species_Shapirotest", shapiro_species)
add_formatted_sheet(wb_norm, "Ind_Shapirotest", shapiro_ind)
add_formatted_sheet(wb_norm, "Rep_Shapirotest", shapiro_rep)
saveWorkbook(wb_norm, file = file.path(out_dir_eda, "normality_tests.xlsx"),
             overwrite = TRUE)

```

\newpage

### Correlation among morphological variables

```{r Check correlation among variables}

# Create an object containing only all morphological variables
morpho_data_stats <- spores_data[,c(5,6,7,8,9)]

# Compute correlation matrix (using pairwise complete obs)
correlation_matrix <- as.data.frame(cor(morpho_data_stats, use = "complete.obs"))

# Create workbook
wb_corr <- createWorkbook()

# Add formatted sheets using existing add_formatted_sheet() function
add_formatted_sheet(wb_corr, "correlation_matrix", correlation_matrix)

# Save excel file with the results
saveWorkbook(wb_corr, file = file.path(out_dir_eda, "correlation_tests.xlsx"),
             overwrite = TRUE)

```

\newpage

## Statistical analysis

### GLOBAL PERMANOVA 

```{r PERMANOVA analysis for E. festucae}

# Select only the species with n >= 2 for conducting the statistical analyses
isolates_per_species <- spores_data_ingroup %>%
  distinct(Species, Ind) %>%
  count(Species, name = "n_isolates") %>%
  arrange(Species)

valid_species <- isolates_per_species %>%
  filter(n_isolates >= 2) %>%
  pull(Species)

# Create data sets including only species with two or more isolates for conducting 
# the statistics
spores_pairwise <- spores_data_ingroup %>%
  filter(Species %in% valid_species) %>%
  droplevels()

plate_means_pairwise <- plate_means_ingroup %>%
  filter(Species %in% valid_species) %>%
  droplevels()

ind_means_pairwise <- ind_means_ingroup %>%
  filter(Species %in% valid_species) %>%
  droplevels()

# Scale the data sets before conducting the analyses
X_spores<- spores_pairwise %>% dplyr::select(all_of(morpho_vars_permpca))
X_spores_scaled <- scale(X_spores)
spores_pairwise_scaled <- spores_pairwise %>%
  dplyr::select(Species, Ind) %>%
  bind_cols(as.data.frame(X_spores_scaled))

X_plate <- plate_means_pairwise %>% dplyr::select(all_of(morpho_vars_permpca))
X_plate_scaled <- scale(X_plate)
plate_means_pairwise_scaled <- plate_means_pairwise %>%
  dplyr::select(Species, Ind) %>%
  bind_cols(as.data.frame(X_plate_scaled))

X_ind <- ind_means_pairwise %>% dplyr::select(all_of(morpho_vars_permpca))
X_ind_scaled <- scale(X_ind)
ind_means_pairwise_scaled <- ind_means_pairwise %>%
  dplyr::select(Species, Ind) %>%
  bind_cols(as.data.frame(X_ind_scaled))

# Summarize n for the three data sets that are going to be included
n_summary <- data.frame(
  Level = c("Spore", "Plate", "Isolate"),
  N_rows = c(nrow(spores_pairwise), nrow(plate_means_pairwise), 
             nrow(ind_means_pairwise)),
  N_species = c(n_distinct(spores_pairwise$Species),
                n_distinct(plate_means_pairwise$Species),
                n_distinct(ind_means_pairwise$Species)),
  N_isolates = c(n_distinct(spores_pairwise$Ind),
                 n_distinct(plate_means_pairwise$Ind),
                 n_distinct(ind_means_pairwise$Ind)))
print(n_summary)


# Create function PERMANOVA + PERMDISP (using standardized traits)
run_permanova_level <- function(df, level_name, response_vars, strata_var = NULL) {
  df <- df %>%
    dplyr::select(all_of(c(response_vars, "Species", strata_var))) %>%
    tidyr::drop_na()
  X  <- df %>% dplyr::select(all_of(response_vars))
  Xs <- scale(X)
  grp <- df$Species
  perm <- adonis2( Xs ~ grp, permutations = 10000, method = "euclidean",
    strata = if(!is.null(strata_var)) df[[strata_var]] else NULL)
  disp <- betadisper(dist(Xs), grp)
  disp_test <- permutest(disp, permutations = 10000)
  list(level = level_name,
       N = nrow(df),
       n_groups = nlevels(grp),
       permanova = perm,
       permdisp = disp_test)
}

# Set seed to improve reproducibility
set.seed(8888)

# Run PERMANOVA at the three levels
res_spore <- run_permanova_level(spores_pairwise_scaled, "Spore-level (exploratory)", 
                                 morpho_vars_permpca, strata_var = "Ind")
res_plate <- run_permanova_level(plate_means_pairwise_scaled, "Plate-level (exploratory)", 
                                 morpho_vars_permpca, strata_var = "Ind")
res_isolate <- run_permanova_level(ind_means_pairwise_scaled, "Isolate-level (Main result)", 
                                   morpho_vars_permpca, strata_var = NULL)

# Print full outputs (GLOBAL PERMANOVA + PERMDISP)
cat("\n\n===== SPORE-LEVEL =====\n")
print(res_spore$permanova)
cat("\nPERMDISP:\n")
print(res_spore$permdisp)

cat("\n\n===== PLATE-LEVEL =====\n")
print(res_plate$permanova)
cat("\nPERMDISP:\n")
print(res_plate$permdisp)

cat("\n\n===== ISOLATE-LEVEL =====\n")
print(res_isolate$permanova)
cat("\nPERMDISP:\n")
print(res_isolate$permdisp)


# Build a compact summary table
extract_perm_summary <- function(res_obj) {
  perm <- res_obj$permanova
  disp <- res_obj$permdisp  # ahora es permutest(disp)
  disp_tab <- disp$tab
  f_col <- intersect(c("F", "F value", "F.Model", "F statistic"), colnames(disp_tab))[1]
  p_col <- intersect(c("Pr(>F)", "Pr(>F)", "p", "p.value"), colnames(disp_tab))[1]
  data.frame(Level = res_obj$level,
             N = res_obj$N,
             Groups = res_obj$n_groups,
             Df_model = perm$Df[1],
             Df_resid = perm$Df[2],
             R2 = perm$R2[1],
             F = perm$F[1],
             p_value = perm$`Pr(>F)`[1],
             PERMDISP_F = if (!is.na(f_col)) disp_tab[1, f_col] else NA_real_,
             PERMDISP_p = if (!is.na(p_col)) disp_tab[1, p_col] else NA_real_,
             row.names = NULL)
}

perm_summary <- dplyr::bind_rows(extract_perm_summary(res_spore),
                                 extract_perm_summary(res_plate),
                                 extract_perm_summary(res_isolate))

# Save outputs of the main result (valid statistical inference)
out_dir_stats <- "G:/ESPORAS_SPCOMP/outputs_script1/1_Spores_dataset/2_STAT"
dir.create(out_dir_stats, recursive = TRUE, showWarnings = FALSE)
write.csv2(perm_summary, file.path(out_dir_stats, "GLOBALPERM_summary.csv"), row.names = FALSE)
sink("G:/ESPORAS_SPCOMP/outputs_script1/1_Spores_dataset/2_STAT/GLOBALPERM_model.txt", split = FALSE)
print(res_isolate$permanova)
sink()

```

\newpage

### PAIRWISE PERMUTATION TEST (isolate level)

```{r Pairwise permutation tests}

# Create the function for pairwise PERMANOVA comparisons
pairwise_adonis2 <- function(df_scaled, response_vars, group_var = "Species",
                             permutations = 10000, method = "euclidean",
                             p_adj = "holm", check_dispersion = TRUE) {

  df_scaled[[group_var]] <- droplevels(as.factor(df_scaled[[group_var]]))
  lvls <- levels(df_scaled[[group_var]])
  pairs <- combn(lvls, 2, simplify = FALSE)
  
  out <- lapply(pairs, function(p) {
    sub <- df_scaled %>% dplyr::filter(.data[[group_var]] %in% p)
    X <- sub %>% dplyr::select(all_of(response_vars))
    g <- droplevels(sub[[group_var]])
    D <- dist(X, method = "euclidean")
    perm <- adonis2(D ~ g, permutations = permutations)
    ns <- table(g)
    n1 <- unname(ns[1]); n2 <- unname(ns[2])
    disp_p <- NA_real_
    disp_F <- NA_real_
    if (check_dispersion) {
      bd <- betadisper(D, g)
      bt <- permutest(bd, permutations = permutations)
      disp_F <- bt$tab[1, "F"]
      disp_p <- bt$tab[1, "Pr(>F)"]
    }

    data.frame(Comparison = paste(p[1], "vs", p[2]),
               n1 = n1, n2 = n2,
               Df = perm$Df[1],
               R2 = perm$R2[1],
               F = perm$F[1],
               p_value = perm$`Pr(>F)`[1],
               PERMDISP_F = disp_F,
               PERMDISP_p = disp_p,
               stringsAsFactors = FALSE)
  })

  dplyr::bind_rows(out) %>%
    dplyr::mutate(p_adj = p.adjust(p_value, method = p_adj)) %>%
    dplyr::arrange(p_adj)
}

# Set seed to impreove reproducibility
set.seed(8888)

# conduct the test at isolate level
pairwise_res <- pairwise_adonis2(df_scaled = ind_means_pairwise_scaled,
                                 response_vars = morpho_vars_permpca)
print(pairwise_res)

# Save the results
write.csv2(pairwise_res,
           file = "G:/ESPORAS_SPCOMP/outputs_script1/1_Spores_dataset/2_STAT/PAIRPERM_summary.csv",
           row.names = FALSE)

```

\newpage

# LLMs for morphological variables (univariate tests)

```{r}

# Set variables and reordering the variables before conducting the analyses and 
# generating the output file
ctrl <- lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
morpho_vars_stats <- c("conidL", "conidW", "conidA", "conidiophL", "conidiophW")
results <- vector("list", length(morpho_vars_stats))
names(results) <- morpho_vars_stats

# Conduct the analyses and generate output file
sink("G:/ESPORAS_SPCOMP/outputs_script1/1_Spores_dataset/2_STAT/LMM_results.txt", 
     split = FALSE)

cat("===================\n")
cat("Linear Mixed Models\n")
cat("===================\n\n")

for (v in morpho_vars_stats) {

  cat("\n=====================\n")
  cat("Variable:", v, "\n")
  cat("=====================\n\n")

  f_full <- as.formula(paste0(v, " ~ Species + (1|Ind/Rep)"))
  f_red  <- as.formula(paste0(v, " ~ Species + (1|Ind)"))

  # Fit full model
  m_full <- lmer(f_full, data = spores_pairwise, REML = TRUE, control = ctrl)
  sing_full <- isSingular(m_full, tol = 1e-5)

  # Choose final model
  if (!sing_full) {
    m_final <- m_full
    f_final <- f_full
    model_label <- "FULL MODEL"
  } else {
    m_final <- lmer(f_red, data = spores_pairwise, REML = TRUE, control = ctrl)
    f_final <- f_red
    model_label <- "REDUCED MODEL (full model singular)"
  }

  sing_final <- isSingular(m_final, tol = 1e-5)

  cat("Selected model (R code):\n")
  cat("m <- lmer(", deparse(f_final), ",\n",
      "          data = spores_pairwise, REML = TRUE, control = ctrl)\n",
      sep = "")
  cat("Model type:", model_label, "\n")
  cat("isSingular(full)  =", sing_full, "\n")
  cat("isSingular(final) =", sing_final, "\n\n")
  cat("\n")

  cat("----- MODEL SUMMARY -----\n")
  print(summary(m_final))
  cat("\n\n\n")

  cat("----- MODEL COMPARISON (Null vs Final; LRT) -----\n")
  m_null <- tryCatch({
    
    # Build a null model keeping the SAME random structure as the final model
    rand_terms <- findbars(formula(m_final))
    rand_str <- if (length(rand_terms) > 0) {
      paste(vapply(rand_terms, function(x) paste0("(", deparse(x), ")"), ""), 
            collapse = " + ")
    } else {
      ""
    }
    f_null <- if (nzchar(rand_str)) {
      as.formula(paste0(v, " ~ 1 + ", rand_str))
    } else {
      as.formula(paste0(v, " ~ 1"))
    }
    lmer(f_null, data = spores_pairwise, REML = FALSE, control = ctrl)
  }, error = function(e) e)

  lrt_null <- tryCatch({
    if (inherits(m_null, "error")) stop(m_null$message)
    anova(m_null, update(m_final, REML = FALSE))
  }, error = function(e) e)

  if (inherits(lrt_null, "error")) {
    cat("Null vs Final comparison failed:\n", lrt_null$message, "\n")
  } else {
    print(lrt_null)
  }
  cat("\n\n\n")

  cat("----- MODEL R2 (marginal / conditional) -----\n")
  r2_vals <- performance::r2(m_final)
  print(r2_vals)

  cat("\n\n\n")

  cat("----- ASSUMPTION CHECKS (model performance) -----\n")
  checks <- tryCatch({
    list(heteroscedasticity = performance::check_heteroscedasticity(m_final),
         outliers = performance::check_outliers(m_final))
  }, error = function(e) e)

  if (inherits(checks, "error")) {
    cat("performance checks failed:\n", checks$message, "\n")
  } else {
  cat("[Heteroscedasticity]\n")
  print(checks$heteroscedasticity)
  cat("\n[Outliers]\n")
  print(checks$outliers)
  }
  cat("\n\n\n")

  cat("----- RESIDUAL NORMALITY (Shapiro-Wilk) -----\n")
  sh <- tryCatch(shapiro.test(resid(m_final)), error = function(e) e)
  if (inherits(sh, "error")) {
    cat("Shapiro test failed:\n", sh$message, "\n")
  } else {
    print(sh)
  }
  cat("\n\n\n")

  # Compute emmeans ONCE, but print later (after pairwise)
  emm <- tryCatch(emmeans(m_final, ~ Species, lmer.df = "kenward-roger"),
                  error = function(e) e)
  
    cat("----- ANOVA TEST (Kenward–Roger) -----\n")
  aov_kr <- tryCatch(anova(m_final, ddf = "Kenward-Roger"),
                     error = function(e) e)
  if (inherits(aov_kr, "error")) {
    cat("Kenward–Roger failed:\n", aov_kr$message, "\n")
  } else {
    print(aov_kr)
  }
  cat("\n\n\n")

   cat("----- PAIRWISE COMPARISONS (Tukey-adjusted) -----\n")
  if (inherits(emm, "error")) {
    cat("emmeans failed (so pairwise cannot be computed):\n", emm$message, "\n")
    pw <- emm
  } else {
    pw <- tryCatch(pairs(emm, adjust = "tukey"),
                   error = function(e) e)
    if (inherits(pw, "error")) {
      cat("Tukey test failed:\n", pw$message, "\n")
    } else {
      print(pw)
    }
  }
  cat("\n\n\n")

    cat("----- COMPACT LETTER DISPLAY -----\n")
  cld_res <- tryCatch({
    if (inherits(emm, "error")) stop("emmeans failed; cannot compute compact letter display.")

    cld_out <- cld(
      emm,
      alpha = 0.05,
      Letters = letters,
      sort = FALSE
    )

    as.data.frame(cld_out)
  }, error = function(e) e)

  if (inherits(cld_res, "error")) {
    cat("Compact letter display failed:\n", cld_res$message, "\n")
  } else {
    print(cld_res)
  }
  cat("\n\n\n")
  
    cat("----- EFFECT SIZES (Hedges g) -----\n")
  eff <- tryCatch({
    if (inherits(emm, "error")) stop("emmeans failed; cannot compute effect sizes.")
    eff_size(emm, sigma = sigma(m_final), edf = df.residual(m_final))
  }, error = function(e) e)

  if (inherits(eff, "error")) {
    cat("Effect sizes failed:\n", eff$message, "\n")
  } else {
    print(eff)
  }
  cat("\n\n\n")

  cat("----- VARIANCE COMPONENTS & PROPORTIONS -----\n\n")

  vc_df <- as.data.frame(VarCorr(m_final))

  # Keep only group + variance, and aggregate (safe against duplicates)
  vc2 <- aggregate(vcov ~ grp, data = vc_df, sum)

  # Ensure Residual exists (VarCorr sometimes omits it)
  if (!"Residual" %in% vc2$grp) {
   vc2 <- rbind(vc2, data.frame(grp = "Residual", vcov = sigma(m_final)^2))
  }

  # Desired order
  desired_order <- c("Ind", "Ind:Rep", "Residual")

  # Keep unexpected levels (don’t convert them to NA)
  extra_levels <- setdiff(unique(as.character(vc2$grp)), desired_order)
  level_order  <- c(desired_order, extra_levels)

  vc2$grp <- factor(as.character(vc2$grp), levels = level_order)
  vc2 <- vc2[order(vc2$grp), , drop = FALSE]

  # compute proportions (THIS WAS MISSING)
  vc2$Proportion <- vc2$vcov / sum(vc2$vcov)

  # Print table
  grp_chr <- as.character(vc2$grp)
  w <- max(nchar(grp_chr), na.rm = TRUE) + 2
  if (!is.finite(w) || w < 7) w <- 10

  fmt_header <- paste0("%-", w, "s %12s %12s\n")
  fmt_row    <- paste0("%-", w, "s %12.6f %12.3f\n")

  cat(sprintf(fmt_header, "Level", "Variance", "Proportion"))
  cat(sprintf(fmt_header, "-----", "--------", "----------"))

  for (i in seq_len(nrow(vc2))) {
    cat(sprintf(fmt_row, grp_chr[i], vc2$vcov[i], vc2$Proportion[i]))
  }

  cat("\n\n\n\n")

  
  # Save diagnostic plots into .jpg files 
  # Helper: safe filename
  safe_name <- function(x) gsub("[^A-Za-z0-9_\\-]+", "_", x)

  # Choose image format: "png" or "jpg"
  img_format <- "jpg"   # you asked for .jpg name

  # Image size (pixels). You can change these.
  img_width  <- 1600
  img_height <- 1200
  img_res    <- 200

  # Function to save diagnostics
  save_lmm_diagnostics <- function(m_final, var_name, out_dir, format = "jpg",
                                   width = 1600, height = 1200, res = 200) {

    # Filename per variable (no overwrite)
    file_path <- file.path(out_dir_stats, paste0(safe_name(var_name), 
                                                 "_LMM_plots.", format))

    # Open device
    if (tolower(format) == "png") {
      png(filename = file_path, width = width, height = height, res = res)
    } else if (tolower(format) %in% c("jpg", "jpeg")) {
      jpeg(filename = file_path, width = width, height = height, quality = 95, 
           res = res)
    } else {
      stop("Unsupported format. Use 'png' or 'jpg'.")
    }

    # Always close device even if an error happens
    old_par <- par(no.readonly = TRUE)
    on.exit({
      par(old_par)
      dev.off()
    }, add = TRUE)

    # Diagnostics data
    resids <- resid(m_final)
    fits   <- fitted(m_final)

    # 2x2 panel
    par(mfrow = c(2, 2), mar = c(4.5, 4.5, 3, 1))

    # 1) Residuals vs Fitted
    plot(fits, resids,
         xlab = "Fitted values",
         ylab = "Residuals",
         main = paste0(var_name, ": Residuals vs Fitted"))
    abline(h = 0, lty = 2)

    # 2) Normal Q-Q
    qqnorm(resids, main = paste0(var_name, ": Normal Q-Q"))
    qqline(resids, lty = 2)

    # 3) Histogram of residuals
    hist(resids, breaks = 25,
         main = paste0(var_name, ": Residuals histogram"),
         xlab = "Residuals")

    # 4) Scale-Location (sqrt(|residuals|) vs fitted)
    plot(fits, sqrt(abs(resids)),
         xlab = "Fitted values",
         ylab = "sqrt(|Residuals|)",
         main = paste0(var_name, ": Scale-Location"))
    lines(lowess(fits, sqrt(abs(resids))), lty = 2)
  }

  # Run and save diagnostics for this variable
  tryCatch({
    save_lmm_diagnostics(m_final, v, out_dir_stats,
                         format = img_format,
                         width = img_width, height = img_height, res = img_res)
  }, error = function(e) {
    cat("Diagnostics plot failed for", v, ":", e$message, "\n")
  })


  # Store results in memory (AFTER objects exist)
  results[[v]] <- list(
    formula_full = f_full,
    formula_final = f_final,
    model_full = m_full,
    model_final = m_final,
    singular_full = sing_full,
    singular_final = sing_final,
    anova_KR = aov_kr,
    emmeans = emm,
    cld = cld_res
  )
}

cat("========================\n")
cat("End of LMM analysis\n")
cat("========================\n")

sink()

```


```{r LMMs using Log transformation (sensitivity)}

# ============================================================
# SENSITIVITY ANALYSIS: log-scale LMM (ONLY for selected vars)
# ============================================================

vars_log_sensitivity <- c("conidW", "conidiophL")

sink("G:/ESPORAS_SPCOMP/outputs_script1/1_Spores_dataset/2_STAT/LMM_log_results.txt", split = TRUE)


cat("=============================\n")
cat("Log-scale sensitivity (LMM)\n")
cat("=============================\n\n")

for (v in vars_log_sensitivity) {

  cat("\n==============================\n")
  cat("Variable (log):", v, "\n")
  cat("==============================\n\n")

  v_log <- paste0("log_", v)

  # Work on a fresh copy per variable (avoids state contamination)
  spores_pairwise_log <- spores_pairwise

  # Prepare safe log transform (handles zeros/negatives)
  x <- spores_pairwise_log[[v]]
  min_x <- suppressWarnings(min(x, na.rm = TRUE))

  if (!is.finite(min_x)) {
    cat("Log sensitivity skipped: variable has no finite values.\n\n\n")
    next
  }

  if (min_x <= 0) {
    offset <- abs(min_x) + 1e-6
    cat(sprintf("Transform used: %s = log(%s + %.6g)  [min(%s)=%.6f]\n\n",
                v_log, v, offset, v, min_x))
    spores_pairwise_log[[v_log]] <- log(spores_pairwise_log[[v]] + offset)
  } else {
    cat(sprintf("Transform used: %s = log(%s)  [min(%s)=%.6f]\n\n",
                v_log, v, v, min_x))
    spores_pairwise_log[[v_log]] <- log(spores_pairwise_log[[v]])
  }

  # Define full/reduced formulas on log-scale
  f_full_log <- as.formula(paste0(v_log, " ~ Species + (1|Ind/Rep)"))
  f_red_log  <- as.formula(paste0(v_log, " ~ Species + (1|Ind)"))

  # Fit full model (log)
  m_full_log <- tryCatch(
    lmer(f_full_log, data = spores_pairwise_log, REML = TRUE,
         control = ctrl, na.action = na.omit),
    error = function(e) e
  )

  if (inherits(m_full_log, "error")) {
    cat("Log sensitivity failed (full model):\n", m_full_log$message, "\n\n\n")
    next
  }

  sing_full_log <- isSingular(m_full_log, tol = 1e-5)

  # Choose final model (log) mirroring your rule
  if (!sing_full_log) {
    m_final_log <- m_full_log
    f_final_log <- f_full_log
    model_label_log <- "FULL MODEL (log)"
  } else {
    m_final_log <- tryCatch(
      lmer(f_red_log, data = spores_pairwise_log, REML = TRUE,
           control = ctrl, na.action = na.omit),
      error = function(e) e
    )
    f_final_log <- f_red_log
    model_label_log <- "REDUCED MODEL (log; full singular)"
  }

  if (inherits(m_final_log, "error")) {
    cat("Log sensitivity failed (final model):\n", m_final_log$message, "\n\n\n")
    next
  }

  sing_final_log <- isSingular(m_final_log, tol = 1e-5)

  cat("Selected log model (R code):\n")
  cat("m_log <- lmer(", deparse(f_final_log), ",\n",
      "               data = spores_pairwise_log, REML = TRUE,\n",
      "               control = ctrl, na.action = na.omit)\n",
      sep = "")
  cat("Model type:", model_label_log, "\n")
  cat("isSingular(full_log)  =", sing_full_log, "\n")
  cat("isSingular(final_log) =", sing_final_log, "\n\n")

  cat("----- LOG MODEL SUMMARY -----\n")
  print(summary(m_final_log))
  cat("\n\n")

  # LRT null vs final (log)
  cat("----- LOG MODEL COMPARISON (Null vs Final; LRT) -----\n")

  m_null_log <- tryCatch({

    rand_terms <- lme4::findbars(formula(m_final_log))
    rand_str <- if (length(rand_terms) > 0) {
      paste(vapply(rand_terms, function(x) paste0("(", deparse(x), ")"), ""),
            collapse = " + ")
    } else {
      ""
    }

    f_null_log <- if (nzchar(rand_str)) {
      as.formula(paste0(v_log, " ~ 1 + ", rand_str))
    } else {
      as.formula(paste0(v_log, " ~ 1"))
    }

    lmer(f_null_log, data = spores_pairwise_log, REML = FALSE,
         control = ctrl, na.action = na.omit)

  }, error = function(e) e)

  lrt_log <- tryCatch({
    if (inherits(m_null_log, "error")) stop(m_null_log$message)
    anova(m_null_log, update(m_final_log, REML = FALSE))
  }, error = function(e) e)

  if (inherits(lrt_log, "error")) {
    cat("Null vs Final (log) comparison failed:\n", lrt_log$message, "\n")
  } else {
    print(lrt_log)
  }
  cat("\n\n")

 
  cat("----- LOG MODEL R2 (marginal / conditional) -----\n")
  r2_log <- tryCatch(performance::r2(m_final_log), error = function(e) e)
  if (inherits(r2_log, "error")) {
    cat("R2 (log) failed:\n", r2_log$message, "\n")
  } else {
    print(r2_log)
  }
  cat("\n\n")

 
  cat("----- LOG ASSUMPTION CHECKS (model performance) -----\n")
  checks_log <- tryCatch({
    list(
      heteroscedasticity = performance::check_heteroscedasticity(m_final_log),
      outliers           = performance::check_outliers(m_final_log)
    )
  }, error = function(e) e)

  if (inherits(checks_log, "error")) {
    cat("performance checks (log) failed:\n", checks_log$message, "\n")
  } else {
    cat("[Heteroscedasticity]\n")
    print(checks_log$heteroscedasticity)
    cat("\n[Outliers]\n")
    print(checks_log$outliers)
  }
  cat("\n\n")

  
  cat("----- LOG RESIDUAL NORMALITY (Shapiro-Wilk) -----\n\n")
  sh_log <- tryCatch(shapiro.test(resid(m_final_log)), error = function(e) e)
  if (inherits(sh_log, "error")) {
    cat("Shapiro (log) failed:\n", sh_log$message, "\n")
  } else {
    print(sh_log)
  }
  cat("\n\n")

  
  cat("----- LOG ANOVA TEST (Kenward–Roger) -----\n")
  aov_kr_log <- tryCatch(anova(m_final_log, ddf = "Kenward-Roger"),
                        error = function(e) e)
  if (inherits(aov_kr_log, "error")) {
    cat("Kenward–Roger (log) failed:\n", aov_kr_log$message, "\n")
  } else {
    print(aov_kr_log)
  }
  cat("\n\n")

  # Extract emmeans + pairwise (log)
  emm_log <- tryCatch(emmeans(m_final_log, ~ Species, 
                              lmer.df = "kenward-roger"), error = function(e) e)

cat("----- LOG PAIRWISE COMPARISONS (Tukey-adjusted) -----\n")
  if (inherits(emm_log, "error")) {
    cat("emmeans failed (so pairwise cannot be computed):\n", emm_log$message, "\n")
    pw_log <- emm_log
  } else {
    pw_log <- tryCatch(pairs(emm_log, adjust = "tukey"),
                   error = function(e) e)
    if (inherits(pw_log, "error")) {
      cat("Tukey test failed:\n", pw_log$message, "\n")
    } else {
      print(pw_log)
    }
  }
  cat("\n\n\n")

    cat("----- LOG COMPACT LETTER DISPLAY -----\n")
  cld_res_log <- tryCatch({
    if (inherits(emm_log, "error")) stop("emmeans failed; cannot compute compact letter display.")

    cld_out_log <- cld(
      emm_log,
      alpha = 0.05,
      Letters = letters,
      sort = FALSE
    )

    as.data.frame(cld_out_log)
  }, error = function(e) e)

  if (inherits(cld_res_log, "error")) {
    cat("Compact letter display failed:\n", cld_res_log$message, "\n")
  } else {
    print(cld_res_log)
  }
  cat("\n\n\n")
  
    cat("----- LOG EFFECT SIZES (Hedges g) -----\n")
  eff_log <- tryCatch({
    if (inherits(emm_log, "error")) stop("emmeans failed; cannot compute effect sizes.")
    eff_size(emm_log, sigma = sigma(m_final_log), edf = df.residual(m_final_log))
  }, error = function(e) e)

  if (inherits(eff_log, "error")) {
    cat("Effect sizes failed:\n", eff_log$message, "\n")
  } else {
    print(eff_log)
  }
  cat("\n\n\n")

  cat("----- LOG ESTIMATED MARGINAL MEANS (emmeans) -----\n")
  if (inherits(emm_log, "error")) {
    cat("emmeans (log) failed:\n", emm_log$message, "\n")
  } else {
    print(emm_log)
  }
  cat("\n\n\n")
}

cat("========================\n")
cat("End of log sensitivity\n")
cat("========================\n")

sink()

```

## Complementary multivariate analyses

### Principal Component Analysis (PCA)

```{r PCA analyses considering all datasets}

# ============================================================
# PCA REPORTING (TXT ONLY) - aligned columns + rounded to 3 decimals
# ============================================================

out_dir_pca <- "G:/ESPORAS_SPCOMP/outputs_script1/1_Spores_dataset/3_PCA"
dir.create(out_dir_pca, recursive = TRUE, showWarnings = FALSE)

# ------------------------------------------------------------
# Helper: pretty aligned table printer (tab-separated, fixed widths)
# - digits_num controls rounding for numeric columns
# ------------------------------------------------------------
print_aligned_table <- function(df, digits_num = 3) {

  if (is.null(df) || nrow(df) == 0) {
    cat("(empty)\n")
    return(invisible(NULL))
  }

  df2 <- df

  # Round numeric columns
  for (j in seq_along(df2)) {
    if (is.numeric(df2[[j]])) df2[[j]] <- round(df2[[j]], digits_num)
  }

  # Convert everything to character (keep NA as "")
  dfc <- as.data.frame(lapply(df2, function(x) {
    x <- as.character(x)
    x[is.na(x)] <- ""
    x
  }), stringsAsFactors = FALSE)

  # Column widths
  widths <- vapply(names(dfc), function(nm) {
    max(nchar(nm), max(nchar(dfc[[nm]]), na.rm = TRUE))
  }, integer(1))

  # Header
  header <- paste0(
    vapply(seq_along(widths), function(i) {
      sprintf(paste0("%-", widths[i], "s"), names(dfc)[i])
    }, character(1)),
    collapse = "  "
  )
  cat(header, "\n", sep = "")

  # Separator
  sep_line <- paste0(
    vapply(seq_along(widths), function(i) {
      paste(rep("-", widths[i]), collapse = "")
    }, character(1)),
    collapse = "  "
  )
  cat(sep_line, "\n", sep = "")

  # Rows
  for (r in seq_len(nrow(dfc))) {
    line <- paste0(
      vapply(seq_along(widths), function(i) {
        sprintf(paste0("%-", widths[i], "s"), dfc[r, i])
      }, character(1)),
      collapse = "  "
    )
    cat(line, "\n", sep = "")
  }
}

# ------------------------------------------------------------
# PCA runner (NO CSVs; PCA object only)
# ------------------------------------------------------------
run_pca <- function(df, vars, file_prefix, out_dir) {

  X <- df %>%
    dplyr::select(all_of(vars)) %>%
    dplyr::mutate(dplyr::across(dplyr::everything(), as.numeric)) %>%
    scale() %>%
    as.data.frame()

  pca <- FactoMineR::PCA(X, graph = FALSE)

  return(pca)
}

# ------------------------------------------------------------
# TXT REPORT FUNCTION
# ------------------------------------------------------------
write_pca_report <- function(pca_obj, df_used, vars_used, pca_name, digits_num = 3) {

  cat("\n\n============================================================\n")
  cat("PCA:", pca_name, "\n")
  cat("============================================================\n\n")

  # ---------------- EIGENVALUES ----------------
  cat("\n---------------------------------------------\n")
  cat("EIGENVALUES & EXPLAINED VARIANCE\n")
  cat("---------------------------------------------\n\n")

  eig <- factoextra::get_eigenvalue(pca_obj)
  eig <- as.data.frame(eig)
  print_aligned_table(eig, digits_num = digits_num)
  cat("\n\n")

  # ---------------- VARIABLE LOADINGS / CONTRIB / COS2 ----------------
  cat("\n---------------------------------------------\n")
  cat("VARIABLE LOADINGS, CONTRIBUTIONS, COS2\n")
  cat("---------------------------------------------\n\n")

  # We compute from standardized data + PC scores (robust & consistent with your output)
  X_scaled <- df_used %>%
    dplyr::select(all_of(vars_used)) %>%
    dplyr::mutate(dplyr::across(dplyr::everything(), as.numeric)) %>%
    scale() %>%
    as.matrix()

  pcs <- as.matrix(pca_obj$ind$coord)
  k <- min(4, ncol(pcs))   # up to PC4 (or less if PCA gives fewer)
  pcs_k <- pcs[, 1:k, drop = FALSE]

  # Remove any constant/NA-only variables (safety)
  keep <- apply(X_scaled, 2, function(x) {
    vv <- var(x, na.rm = TRUE)
    is.finite(vv) && vv > 0
  })
  X_scaled <- X_scaled[, keep, drop = FALSE]

  if (ncol(X_scaled) == 0) {
    cat("WARNING: no non-constant variables available.\n\n")
  } else {

    # Loadings interpreted as correlations variable–PC
    coord_mat <- cor(X_scaled, pcs_k, use = "pairwise.complete.obs")

    # Eigenvalues for first k components
    eigvals <- pca_obj$eig[1:k, 1]

    # Contributions (%): 100 * coord^2 / eigenvalue
    contrib_mat <- sweep(coord_mat^2, 2, eigvals, FUN = "/") * 100

    # cos2 per variable: coord^2 / sum(coord^2 across included PCs)
    denom <- rowSums(coord_mat^2)
    denom[denom == 0] <- NA_real_
    cos2_mat <- (coord_mat^2) / denom

    loadings <- as.data.frame(coord_mat) %>% tibble::rownames_to_column("Variable")
    contrib  <- as.data.frame(contrib_mat) %>% tibble::rownames_to_column("Variable")
    cos2     <- as.data.frame(cos2_mat) %>% tibble::rownames_to_column("Variable")

    cat(">> Loadings:\n\n")
    print_aligned_table(loadings, digits_num = digits_num)
    cat("\n\n")

    cat(">> Contributions (%):\n\n")
    print_aligned_table(contrib, digits_num = digits_num)
    cat("\n\n")

    cat(">> cos2:\n\n")
    print_aligned_table(cos2, digits_num = digits_num)
    cat("\n\n")
  }

  # ---------------- INDIVIDUAL SCORES ----------------
  cat("\n---------------------------------------------\n")
  cat("INDIVIDUAL SCORES\n")
  cat("---------------------------------------------\n\n")

  scores <- as.data.frame(pca_obj$ind$coord) %>%
    tibble::rownames_to_column("Individual")

  if ("Species" %in% names(df_used)) {
    scores$Species <- df_used$Species
  }

  # keep first 4 dims if present
  keep_dims <- intersect(c("Dim.1", "Dim.2", "Dim.3", "Dim.4"), names(scores))
  scores_show <- scores[, c("Individual", keep_dims, intersect("Species", names(scores))), drop = FALSE]

  # round numeric dims (for printing)
  for (nm in keep_dims) scores_show[[nm]] <- round(scores_show[[nm]], digits_num)

  cat("NOTE: Only first 20 rows printed.\n\n")
  print_aligned_table(utils::head(scores_show, 20), digits_num = digits_num)
  cat("\n\n")

  # ---------------- ANOVA PC1/PC2 + Tukey ----------------
  cat("\n---------------------------------------------\n")
  cat("TESTS ON PC AXES (Species)\n")
  cat("---------------------------------------------\n\n")

  if (!("Species" %in% names(scores))) {
    cat("WARNING: No Species column found; skipping ANOVA/Tukey.\n\n")
  } else {

    scores$Species <- factor(scores$Species)

    if ("Dim.1" %in% names(scores)) {
      cat(">> ANOVA PC1\n")
      a1 <- aov(Dim.1 ~ Species, data = scores)
      print(summary(a1))
      cat("\n")

      cat(">> Tukey PC1\n")
      print(TukeyHSD(a1))
      cat("\n\n")
    }

    if ("Dim.2" %in% names(scores)) {
      cat(">> ANOVA PC2\n")
      a2 <- aov(Dim.2 ~ Species, data = scores)
      print(summary(a2))
      cat("\n")

      cat(">> Tukey PC2\n")
      print(TukeyHSD(a2))
      cat("\n\n")
    }
  }

  # ---------------- VARIABLE–PC CORRELATIONS ----------------
  cat("\n---------------------------------------------\n")
  cat("VARIABLE–PC CORRELATIONS\n")
  cat("---------------------------------------------\n\n")

  if (!exists("coord_mat")) {
    cat("NOTE: correlations already printed as 'Loadings' (same matrix).\n\n")
  } else {
    cor_pc <- as.data.frame(coord_mat) %>% tibble::rownames_to_column("Variable")
    print_aligned_table(cor_pc, digits_num = digits_num)
    cat("\n\n")
  }
}

# ------------------------------------------------------------
# NEW DATASETS
# ------------------------------------------------------------
ind_means_Farun <- ind_means %>%
  dplyr::filter(Species != "Ftrich", Species != "Flamb")

# ------------------------------------------------------------
# RUN PCAs
# ------------------------------------------------------------
pca_main <- run_pca(
  df = ind_means_pairwise,
  vars = morpho_vars_permpca,
  file_prefix = "PCA_main_noFarun",
  out_dir = out_dir_pca
)

pca_farun <- run_pca(
  df = ind_means_Farun,
  vars = morpho_vars_permpca,
  file_prefix = "PCA_withFarun",
  out_dir = out_dir_pca
)

# ------------------------------------------------------------
# WRITE TXT REPORT
# ------------------------------------------------------------
sink(file.path(out_dir_pca, "PCA_results.txt"), split = FALSE)

cat("===================\n")
cat("PCA RESULTS\n")
cat("===================\n")

write_pca_report(
  pca_obj = pca_main,
  df_used = ind_means_pairwise,
  vars_used = morpho_vars_permpca,
  pca_name = "PCA_main_noFarun",
  digits_num = 3
)

write_pca_report(
  pca_obj = pca_farun,
  df_used = ind_means_Farun,
  vars_used = morpho_vars_permpca,
  pca_name = "PCA_withFarun",
  digits_num = 3
)

cat("\n========================\n")
cat("End of PCA analysis\n")
cat("========================\n")

sink()


# ------------------------------------------------------------
# PREP OBJECTS FOR PLOTTING (scores + eigenvalues)
# ------------------------------------------------------------
pca_df_main  <- as.data.frame(pca_main$ind$coord)
pca_df_farun <- as.data.frame(pca_farun$ind$coord)

eig_main  <- factoextra::get_eigenvalue(pca_main)
eig_farun <- factoextra::get_eigenvalue(pca_farun)

# Safety checks: group vectors must match rows of PCA coords
stopifnot(nrow(pca_df_main)  == nrow(ind_means_pairwise))
stopifnot(nrow(pca_df_farun) == nrow(ind_means_Farun))



# Plot the PCA analysis
plot_pca_group <- function(data, group_vector, color_palette, title, eig_table) {

  data$Group <- group_vector

  # Extract variances (PC1 y PC2)
  pc1_var <- round(eig_table[1, "variance.percent"], 2)
  pc2_var <- round(eig_table[2, "variance.percent"], 2)

  # Clean rows
  data2 <- data %>%
    dplyr::mutate(Group = as.factor(Group)) %>%
    dplyr::filter(is.finite(Dim.1), is.finite(Dim.2), !is.na(Group))

  # Calculate centroids
  centroids <- data2 %>%
    dplyr::group_by(Group) %>%
    dplyr::summarise(
      Dim.1 = mean(Dim.1, na.rm = TRUE),
      Dim.2 = mean(Dim.2, na.rm = TRUE),
      .groups = "drop"
    )

  # ------------------------------------------------------------
  # MANUAL ELLIPSES (robust; ALWAYS draws if possible)
  # ------------------------------------------------------------
  make_ellipses <- function(df, level = 0.95, n = 200) {

    # radius for chi-square in 2D
    r <- sqrt(stats::qchisq(level, df = 2))
    theta <- seq(0, 2*pi, length.out = n)
    circle <- cbind(cos(theta), sin(theta))

    out <- df %>%
      dplyr::group_by(Group) %>%
      dplyr::group_modify(~{
        d <- .x

        # need >= 3 points for covariance
        if (nrow(d) < 3) return(data.frame())

        mu <- c(mean(d$Dim.1), mean(d$Dim.2))
        S  <- stats::cov(cbind(d$Dim.1, d$Dim.2), use = "pairwise.complete.obs")

        if (any(!is.finite(S))) return(data.frame())

        # ensure positive-definite covariance (nearPD if needed)
        S_pd <- tryCatch({
          as.matrix(Matrix::nearPD(S, corr = FALSE)$mat)
        }, error = function(e) NULL)

        if (is.null(S_pd)) return(data.frame())

        # chol may still fail if too degenerate
        L <- tryCatch(chol(S_pd), error = function(e) NULL)
        if (is.null(L)) return(data.frame())

        pts <- t(mu + r * t(circle %*% L))

        data.frame(
          Dim.1 = pts[, 1],
          Dim.2 = pts[, 2]
        )
      }) %>%
      dplyr::ungroup()

    out
  }

  ellipse_df <- make_ellipses(data2, level = 0.95, n = 200)

  # Create PCA plot
  ggplot2::ggplot(data2, ggplot2::aes(x = Dim.1, y = Dim.2, fill = Group, color = Group)) +
    # Ellipses behind points
    ggplot2::geom_polygon(
      data = ellipse_df,
      ggplot2::aes(x = Dim.1, y = Dim.2, group = Group, fill = Group, color = Group),
      alpha = 0.10,
      linewidth = 0.2,
      inherit.aes = FALSE
    ) +
    ggplot2::geom_point(shape = 21, size = 2, alpha = 0.9) +
    ggplot2::geom_point(
      data = centroids,
      ggplot2::aes(x = Dim.1, y = Dim.2, fill = Group),
      shape = 24, size = 3, color = "black", stroke = 0.4, inherit.aes = FALSE
    ) +
    ggplot2::scale_fill_manual(values = color_palette) +
    ggplot2::scale_color_manual(values = color_palette) +
    ggplot2::labs(
      title = title,
      x = paste0("PC1 (", pc1_var, "%)"),
      y = paste0("PC2 (", pc2_var, "%)")
    ) +
    ggplot2::coord_fixed(ratio = 1) +
    ggplot2::theme_bw(base_size = 8) +
    ggplot2::theme(
      plot.title = ggplot2::element_text(size = 9, hjust = 0.5),
      legend.position = "none",
      axis.title = ggplot2::element_text(size = 7),
      axis.text = ggplot2::element_text(size = 6),
      axis.line = ggplot2::element_line(linewidth = 0.3, color = "black")) +
    ggplot2::theme(panel.grid = element_blank())
}


# Choose palettes
pal_main  <- if ("Farun" %in% ind_means_pairwise$Species) species_col else species_col_ingroup
pal_farun <- if ("Farun" %in% ind_means_Farun$Species) species_col else species_col_ingroup

# Build plots
p_main <- plot_pca_group(
  data = pca_df_main,
  group_vector = ind_means_pairwise$Species,
  color_palette = pal_main,
  title = "PCA (no Farun)",
  eig_table = eig_main
)

p_farun <- plot_pca_group(
  data = pca_df_farun,
  group_vector = ind_means_Farun$Species,
  color_palette = pal_farun,
  title = "PCA (with Farun)",
  eig_table = eig_farun
)

# Save PDFs
ggplot2::ggsave(
  filename = file.path(out_dir_pca, "PCA_ingroup.pdf"),
  plot = p_main,
  width = 6, height = 6, units = "in"
)

ggplot2::ggsave(
  filename = file.path(out_dir_pca, "PCA_allsp.pdf"),
  plot = p_farun,
  width = 6, height = 6, units = "in"
)

```

\newpage

# Analysis of the growth rate of _Epichloë_ cultures on potato dextrose agar (PDA)

## Load, explore & mutate dataset

```{r Data set load, exploration and transformation}
# Set directory for outputs
out_dir_cult <- "G:/ESPORAS_SPCOMP/outputs_script1/2_Culture_growth_dataset/"

# Load dataset into working session
growth_data <- read.table("G:/ESPORAS_SPCOMP/MORPHOLOGY/culture_growth_rate_dataset_final.csv", 
                          dec = ",", sep = ";", header = TRUE, row.names = 1)

# Create new variables
growth_data <- growth_data %>%
  mutate(Rep_ind = paste(Ind, Rep, sep = "-"),
         Day = as.factor(Day),
         Ind = as.factor(Ind),
         Rep = as.factor(Rep),
         Species = as.factor(Species))

```

## Computation of Growth rate Mixed-effects model fitting

```{r Growth rate computation}

# Compute growth rate between D0 and D21 (mm/day)
growth_D0 <- growth_data %>%
  filter(Day == "D0") %>%
  dplyr::select(Rep_ind, Diameter_D0 = Diameter)

growth_D21 <- growth_data %>%
  filter(Day == "D21") %>%
  dplyr::select(Rep_ind, Diameter_D21 = Diameter)

growth_summary <- left_join(growth_D0, growth_D21, by = "Rep_ind") %>%
  mutate(GR = (Diameter_D21 - Diameter_D0) / 21) %>%
  left_join(
    growth_data %>%
      dplyr::select(Rep_ind, Ind, Species, Rep) %>%
      distinct(),
    by = "Rep_ind"
  ) %>%
  filter(!is.na(GR))

# Create species vectors
ord_no_farun <- c("Ftrich", "Flamb", "Fyves", "Froth", "Fnigr", "Fprui")
ord_with_farun <- c("Ftrich", "Flamb", "Fyves", "Froth", "Fnigr", "Fprui", "Farun")

# Datasets preparation
growth_data_plot <- growth_data %>%
  mutate(
    Diameter = as.numeric(Diameter),
    Day_num  = as.numeric(str_remove(Day, "^D")),
    Species  = factor(Species, levels = ord_with_farun),
    Ind      = as.factor(Ind)
  )

growth_mean <- growth_data_plot %>%
  group_by(Species, Ind, Day_num) %>%
  summarise(Diameter = mean(Diameter), .groups = "drop")

species_mean <- growth_mean %>%
  group_by(Species, Day_num) %>%
  summarise(Diameter = mean(Diameter), .groups = "drop")

growth_ind_mean <- growth_summary %>%
  group_by(Species, Ind) %>%
  summarise(GR = mean(GR), .groups = "drop") %>%
  mutate(Species = factor(Species, levels = ord_with_farun))

# Create and save summary table of GR (Species level)
summary_table_GR <- growth_ind_mean %>%
  dplyr::group_by(Species) %>%
  dplyr::summarise(
    N_isolates = dplyr::n(),
    Mean = mean(GR),
    SD = sd(GR),
    SE = SD / sqrt(N_isolates),
    CV = (SD / Mean) * 100,
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    Mean = round(Mean, 2),
    SE   = round(SE, 2),
    CV   = round(CV, 1),
    GR_mean_SE = paste0(Mean, " ± ", SE)
  ) %>%
  dplyr::select(Species, GR_mean_SE, CV)

write.table(
  summary_table_GR,
  file = paste0(out_dir_cult, "summary_table_GR.txt"),
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)

```

```{r Mixed-effects model fitting and diagnostics}

# -------------------------
# MIXED-EFFECTS MODEL (your biological question)
# Species = fixed effect, Isolate (Ind) nested within Species = random effect
# -------------------------
growth_filtered <- subset(growth_summary,
                          !(Species %in% c("Flamb","Ftrich", "Farun")))

model_species <- lmer(GR ~ Species + (1 | Species:Ind), data = growth_filtered)

# -------------------------
# SAVE RESULTS TO TXT
# -------------------------
sink(file.path(out_dir_cult, "lme_results_noFarun.txt"), split = FALSE)

cat(">>> Mixed-effects model: GR ~ Species + (1 | Species:Ind) <<<\n\n")
print(anova(model_species))

cat("\nPost-hoc test (Tukey):\n")
print(emmeans(model_species, pairwise ~ Species, adjust = "tukey"))

cat("\n\n>>> Random-effects variance (Isolate within Species) <<<\n\n")
print(VarCorr(model_species))

# Optional: ICC for isolate-within-species variance
vc <- as.data.frame(VarCorr(model_species))
var_ind <- vc$vcov[vc$grp == "Species:Ind"]
var_res <- vc$vcov[vc$grp == "Residual"]
ICC_ind <- var_ind / (var_ind + var_res)

cat("\n\nICC (proportion of variance due to isolates within species):\n")
print(ICC_ind)

sink()

# -------------------------
# DIAGNOSTIC PLOTS TO PDF
# -------------------------
pdf("G:/ESPORAS_SPCOMP/outputs_script1/2_Culture_growth_dataset/lme_diagnostics_noFarun.pdf", 
    width = 8, height = 10)

par(mfrow = c(2, 2))

plot(resid(model_species), main = "Residuals: Species model")
qqnorm(resid(model_species)); qqline(resid(model_species))

plot(fitted(model_species), resid(model_species),
     main = "Residuals vs Fitted",
     xlab = "Fitted values", ylab = "Residuals")
abline(h = 0)

invisible(dev.off())
par(mfrow = c(1, 1))

```


```{r}

# -------------------------
# GR PLOTS (diamonds with black border + colored fill)
# -------------------------
p_gr_no_farun <- ggplot(filter(growth_ind_mean, Species %in% ord_no_farun),
                        aes(x = Species, y = GR, color = Species)) +
  geom_point(size = 3, alpha = 0.35,
             position = position_jitter(width = 0.05)) +
  stat_summary(fun = mean, geom = "point",
               size = 3.5, shape = 23,
               stroke = 1, lwd = 0.8,
               aes(fill = Species),
               color = "black") +
  scale_color_manual(values = species_col, drop = FALSE) +
  scale_fill_manual(values = species_col, drop = FALSE) +
  labs(x = "Species", y = "Growth rate (mm/day)", title = "Panel 1: without Farun") +
  theme_bw() +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.05)))

p_gr_with_farun <- ggplot(filter(growth_ind_mean, Species %in% ord_with_farun),
                          aes(x = Species, y = GR, color = Species)) +
  geom_point(size = 3, alpha = 0.35,
             position = position_jitter(width = 0.05)) +
  stat_summary(fun = mean, geom = "point",
               size = 3.5, shape = 23,
               stroke = 1, lwd = 0.8,
               aes(fill = Species),
               color = "black") +
  scale_color_manual(values = species_col, drop = FALSE) +
  scale_fill_manual(values = species_col, drop = FALSE) +
  labs(x = "Species", y = "Growth rate (mm/day)", title = "Panel 2: with Farun") +
  theme_bw() +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.05)))
  theme(legend.position = "none")

# -------------------------
# CURVE PLOTS (isolates colored by species, alpha=0.35; mean thicker)
# -------------------------
p_curve_no_farun <- ggplot(filter(growth_mean, Species %in% ord_no_farun),
                           aes(x = Day_num, y = Diameter, group = Ind, color = Species)) +
  geom_line(size = 0.5, alpha = 0.5) +
  geom_line(data = filter(species_mean, Species %in% ord_no_farun),
            inherit.aes = FALSE,
            aes(x = Day_num, y = Diameter, group = Species, color = Species),
            size = 1, alpha = 1) +
  geom_point(data = filter(species_mean, Species %in% ord_no_farun),
             inherit.aes = FALSE,
             aes(x = Day_num, y = Diameter, color = Species),
             size = 2) +
  scale_color_manual(values = species_col, drop = FALSE) +
  scale_x_continuous(breaks = c(0, 2, 7, 14, 21), limits = c(0, 21)) +
  labs(x = "Days after inoculation",
       y = "Colony diameter (mm)",
       title = "Panel 1: without Farun") +
  theme_bw() +
  theme(legend.position = "none")

p_curve_with_farun <- ggplot(filter(growth_mean, Species %in% ord_with_farun),
                             aes(x = Day_num, y = Diameter, group = Ind, color = Species)) +
  geom_line(size = 0.5, alpha = 0.5) +
  geom_line(data = filter(species_mean, Species %in% ord_with_farun),
            inherit.aes = FALSE,
            aes(x = Day_num, y = Diameter, group = Species, color = Species),
            size = 1, alpha = 1) +
  geom_point(data = filter(species_mean, Species %in% ord_with_farun),
             inherit.aes = FALSE,
             aes(x = Day_num, y = Diameter, color = Species),
             size = 2) +
  scale_color_manual(values = species_col, drop = FALSE) +
  scale_x_continuous(breaks = c(0, 2, 7, 14, 21), limits = c(0, 21)) +
  labs(x = "Days after inoculation",
       y = "Colony diameter (mm)",
       title = "Panel 2: with Farun") +
  theme_bw() +
  theme(legend.position = "none")  
  
  
# -------------------------
# COMBINE PANELS
# -------------------------
fig_no_farun  <- p_gr_no_farun + p_curve_no_farun
fig_with_farun <- p_gr_with_farun + p_curve_with_farun

# -------------------------
# SAVE TO PDF
# -------------------------
ggsave(filename = file.path(out_dir_cult, "GR_noFarun.pdf"),
       plot = fig_no_farun, device = cairo_pdf,
       width = 10, height = 4, units = "in")

ggsave(filename = file.path(out_dir_cult, "GR_Farun.pdf"),
       plot = fig_with_farun, device = cairo_pdf,
       width = 10, height = 4, units = "in")

```

```{r End of script}

# Message announcing the end of the script and where to find the output files
cat("\n✅ Script executed successfully.\n\nAll output files are available in 'FrothEpi_Github/Fungal_endophyte/data/outputs_script1 \nwithin the GitHub repository downloaded by the user.\n")


CAMBIA ESTOOOO!!!!!!!
```