---
title: "Construction and basic editing of phylogenetic trees for _Epichloë_ isolates"
author: "Sotomayor-Alge, Alba"
date: "02-06-2026"
output:
  bookdown::pdf_book:
    toc: yes
    toc_depth: 3
    number_sections: yes
    highlight: tango
    latex_engine: xelatex
    keep_tex: false
---

\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
```

# PHYLOGENETIC ANALYSIS

# 1. Initial Setup and Package Loading

```{r Initialize working environment}

# Function to silently load packages without warnings or messages
load_packages <- function(pkgs) {
  invisible(lapply(pkgs, function(pkg) {
      suppressWarnings(suppressMessages(
          if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
            stop(paste("❌ Package not found:", pkg), call. = FALSE)
          }))
    }))
}

# Vector of required packages
packages_needed <- c("dplyr", "tidyr", "ggtree", "ggtreeExtra", "ggplot2", 
                     "treeio", "scatterpie", "ape", "openxlsx", "tibble",
                     "patchwork", "ape", "ggforce", "stringr", "ggtext", "grid",
                     "svglite")

# Call the function to load them
load_packages(packages_needed)

# Print the information of the working session in order to ensure reproducibility
sessionInfo()

```

```{r Create working directories}

# Create directories within the file system to save the results of the analyses
dir.create("../data/outputs_script_2", recursive = TRUE, showWarnings = FALSE)
dir.create("../data/outputs_script_2/1_actG", recursive = TRUE, showWarnings = FALSE)
dir.create("../data/outputs_script_2/2_CalM", recursive = TRUE, showWarnings = FALSE)
dir.create("../data/outputs_script_2/3_ITS", recursive = TRUE, showWarnings = FALSE)
dir.create("../data/outputs_script_2/4_tefA", recursive = TRUE, showWarnings = FALSE)
dir.create("../data/outputs_script_2/5_tubB", recursive = TRUE, showWarnings = FALSE)
dir.create("../data/outputs_script_2/6_MSCtree", recursive = TRUE, showWarnings = FALSE)
dir.create("../data/outputs_script_2/7_MLtree", recursive = TRUE, showWarnings = FALSE)

```

\newpage

# 2. Individual phylogenetic trees

## 2.1 Extract UFbootstrap support values+

```{r Extraction of UFbootstrap Values from Individual Gene trees (5 genes)}

# Function to extract UFbootstrap values from an IQ-TREE tree file and export
# them as an Excel table compatible with ggtree node-based joins
extract_iqtree_ufboot_to_xlsx <- function(treefile, out_xlsx) {

  # Read the phylogenetic tree generated by IQ-TREE
  tr <- read.tree(treefile)

  # Retrieve the number of terminal taxa and internal nodes
  n_tip  <- length(tr$tip.label)
  n_node <- tr$Nnode

  # Create a node-level table using the same internal node numbering as ggtree
  boot <- tibble(node   = (n_tip + 1):(n_tip + n_node),
                 ufboot = suppressWarnings(as.numeric(tr$node.label)))

  # Export the UFbootstrap table as an Excel file
  write.xlsx(boot, out_xlsx, overwrite = TRUE)

  # Report successful file creation
  message("\nExcel created successfully: ", out_xlsx)
}


# Define the set of genes to be processed
genes <- c("actG", "CalM", "ITS", "tefA", "tubB")


# Iterate over each gene and extract its corresponding UFbootstrap values
for (i in seq_along(genes)) {
  gene <- genes[i]
  
  # Select the gene-specific output directory
  out_dir_phy <- paste0("../data/outputs_script_2/", i, "_", gene, "/")

  # Extract UFbootstrap values and save them as a gene-specific Excel file
  extract_iqtree_ufboot_to_xlsx(
    treefile = paste0("../data/Phylogenetic_analysis/TREES/MSA_", 
                      gene, "_UFBoot1000.treefile"),
    out_xlsx = paste0(out_dir_phy, "UFBoot_values_", gene, ".xlsx")
  )
}

```


## 2.2 Format, plot and save phylogenetic trees

```{r Plot and save individual gene phylogenetic trees (5 genes)}


# ------------------------------------------------------------
# 1. DEFINE GENES AND PANEL LABELS
# ------------------------------------------------------------

genes <- c("actG", "CalM", "ITS", "tefA", "tubB")

panel_labels <- c(
  "actG" = "S2A",
  "CalM" = "S2B",
  "ITS" = "S2C",
  "tefA" = "S2D",
  "tubB" = "S2E"
)


# ------------------------------------------------------------
# 2. DEFINE FINAL LABELS FOR OUR SAMPLES
# ------------------------------------------------------------

uz_labels <- c(
  "UZ_11.23_8_" = "UZ_11.23(8) - Froth",
  "UZ_7.23_8_"  = "UZ_7.23(8) - Fyves",
  "UZ_55.22_2_" = "UZ_55.22(2) - Fprui",
  "UZ_9.23_4_"  = "UZ_9.23(4) - Froth",
  "UZ_55.22_4_" = "UZ_55.22(4) - Fprui",
  "UZ_44.22_1_" = "UZ_44.22(1) - Fnigr",
  "UZ_13.22_5_" = "UZ_13.22(5) - Flamb",
  "UZ_4.22_1_"  = "UZ_4.22(1) - Ftrich",
  "UZ_7.23_11_" = "UZ_7.23(11) - Fyves",
  "UZ_44.22_4_" = "UZ_44.22(4) - Fnigr"
)


# ------------------------------------------------------------
# 3. CLEAN RAW TIP LABELS
# ------------------------------------------------------------

clean_tip_label <- function(x) {

  x <- trimws(x)

  # Replace our sample IDs by their final manuscript labels
  is_uz <- x %in% names(uz_labels)
  x[is_uz] <- unname(uz_labels[x[is_uz]])

  # Clean reference labels
  x[!is_uz] <- gsub("_", " ", x[!is_uz])

  # Standardize full genus names
  x[!is_uz] <- gsub("\\bEpichlo[eë]\\b", "E.", x[!is_uz])
  x[!is_uz] <- gsub("\\bClaviceps\\b", "C.", x[!is_uz])

  # Ensure a space after genus abbreviations
  x[!is_uz] <- gsub("^E\\.\\s*", "E. ", x[!is_uz])
  x[!is_uz] <- gsub("^C\\.\\s*", "C. ", x[!is_uz])

  # Standardize Epichloë typhina subspecies labels
  x[!is_uz] <- gsub("^E\\. typhinatyphina\\b",
                    "E. typhina subsp. typhina", x[!is_uz])

  x[!is_uz] <- gsub("^E\\. typhinaclarkii\\b",
                    "E. typhina subsp. clarkii", x[!is_uz])

  x[!is_uz] <- gsub("^E\\. typhinapoae\\b",
                    "E. typhina subsp. poae", x[!is_uz])

  # Normalize spaces
  x <- gsub("\\s+", " ", x)

  trimws(x)
}


# ------------------------------------------------------------
# 4. IDENTIFY OUR SAMPLES AND OUTGROUP
# ------------------------------------------------------------

is_our_sample <- function(x) {
  grepl("^UZ_", x)
}

is_outgroup <- function(x) {
  grepl("^C\\. purpurea\\b", x, ignore.case = TRUE)
}


# ------------------------------------------------------------
# 5. FORMAT TIP LABELS
# ------------------------------------------------------------

format_tip_label <- function(x) {

  # Our samples: complete label in bold
  if (is_our_sample(x)) {
    return(paste0("**", x, "**"))
  }

  # Epichloë typhina subspecies
  if (grepl("^E\\. typhina subsp\\. [a-z-]+\\b", x)) {

    subspecies <- stringr::str_extract(
      x,
      "(?<=subsp\\. )[a-z-]+"
    )

    prefix <- paste0("E. typhina subsp. ", subspecies)

    rest <- trimws(
      sub(
        paste0(
          "^E\\. typhina subsp\\. ",
          subspecies
        ),
        "",
        x
      )
    )

    return(
      paste0(
        "*E. typhina* subsp. *",
        subspecies,
        "*",
        ifelse(rest == "", "", paste0(" ", rest))
      )
    )
  }

  # Standard Epichloë species names
  if (grepl("^E\\. [a-z-]+\\b", x)) {

    species <- stringr::str_extract(
      x,
      "^E\\. [a-z-]+"
    )

    rest <- trimws(
      sub(
        paste0("^", gsub("\\.", "\\\\.", species)),
        "",
        x
      )
    )

    return(
      paste0(
        "*",
        species,
        "*",
        ifelse(rest == "", "", paste0(" ", rest))
      )
    )
  }

  # Claviceps purpurea outgroup
  if (grepl("^C\\. purpurea\\b", x)) {

    rest <- trimws(
      sub("^C\\. purpurea", "", x)
    )

    return(
      paste0(
        "*C. purpurea*",
        ifelse(rest == "", "", paste0(" ", rest))
      )
    )
  }

  x
}


# ------------------------------------------------------------
# 6. FORMAT PANEL TITLES
# ------------------------------------------------------------

format_panel_title <- function(gene_name) {

  panel <- panel_labels[[gene_name]]

  if (gene_name == "ITS") {
    paste0("**", panel, "**  ITS")
  } else {
    paste0("**", panel, "**  *", gene_name, "*")
  }
}


# ------------------------------------------------------------
# 7. GENERATE FINAL TREE
# ------------------------------------------------------------

process_tree_final <- function(gene_name, index) {

  # Input and output paths
  treefile_path <- paste0(
    "../data/Phylogenetic_analysis/TREES/MSA_",
    gene_name,
    "_UFBoot1000.treefile"
  )

  output_dir <- paste0(
    "../data/outputs_script_2/",
    index,
    "_",
    gene_name,
    "/"
  )

  if (!file.exists(treefile_path)) {
    stop("Tree file not found: ", treefile_path)
  }

  if (!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
  }


  # ----------------------------------------------------------
  # READ AND PREPARE TREE
  # ----------------------------------------------------------

  tree <- ape::read.tree(treefile_path)

  # Same branch-length transformation used previously
  tree$edge.length <- sqrt(sqrt(tree$edge.length))

  # Clean tip labels
  tree$tip.label <- vapply(
    tree$tip.label,
    clean_tip_label,
    character(1)
  )


  # ----------------------------------------------------------
  # CREATE BASE TREE
  # ----------------------------------------------------------

  base_tree <- ggtree::ggtree(tree, linewidth = 0.50)

  tree_data <- base_tree$data


  # ----------------------------------------------------------
  # PREPARE TIP LABELS
  # ----------------------------------------------------------

  tip_data <- tree_data %>%
    dplyr::filter(isTip) %>%
    dplyr::mutate(
      display_label = vapply(label, format_tip_label, character(1)),
      sample_type = dplyr::case_when(
        vapply(label, is_our_sample, logical(1)) ~ "sample",
        vapply(label, is_outgroup, logical(1)) ~ "outgroup",
        TRUE ~ "reference"
      )
    )


  # ----------------------------------------------------------
  # PREPARE UFBOOTSTRAP VALUES
  # ----------------------------------------------------------

  node_data <- tree_data %>%
    dplyr::filter(!isTip) %>%
    dplyr::mutate(
      ufboot = suppressWarnings(as.numeric(label))
    ) %>%
    dplyr::filter(!is.na(ufboot), ufboot >= 30)


  # ----------------------------------------------------------
  # DEFINE PLOT COORDINATES
  # ----------------------------------------------------------

  max_tree_x <- max(tree_data$x, na.rm = TRUE)
  max_y <- max(tree_data$y, na.rm = TRUE)

  # Aligned tip-label position
  label_x <- max_tree_x + max_tree_x * 0.20
  label_line_end <- label_x - max_tree_x * 0.015

  # Fixed bootstrap displacement relative to every node
  bootstrap_x_offset <- max_tree_x * 0.035
  bootstrap_y_offset <- 0.32

  node_data <- node_data %>%
    dplyr::mutate(
      bootstrap_x = x - bootstrap_x_offset,
      bootstrap_y = y + bootstrap_y_offset
    )


  # ----------------------------------------------------------
  # CONSTRUCT FINAL TREE
  # ----------------------------------------------------------

  final_tree <- base_tree +

    # Black dotted lines connecting tips and labels
    geom_segment(
      data = tip_data,
      aes(x = x, xend = label_line_end, y = y, yend = y),
      inherit.aes = FALSE,
      colour = "black",
      linewidth = 0.40,
      linetype = "dotted"
    ) +

    # Reference taxa
    ggtext::geom_richtext(
      data = tip_data %>%
        dplyr::filter(sample_type == "reference"),
      aes(x = label_x, y = y, label = display_label),
      inherit.aes = FALSE,
      hjust = 0,
      vjust = 0.5,
      size = 3.8,
      fill = NA,
      label.color = NA,
      label.padding = grid::unit(c(0, 0, 0, 0), "pt"),
      colour = "black"
    ) +

    # Our samples
    ggtext::geom_richtext(
      data = tip_data %>%
        dplyr::filter(sample_type == "sample"),
      aes(x = label_x, y = y, label = display_label),
      inherit.aes = FALSE,
      hjust = 0,
      vjust = 0.5,
      size = 3.8,
      fill = NA,
      label.color = NA,
      label.padding = grid::unit(c(0, 0, 0, 0), "pt"),
      colour = "black"
    ) +

    # Claviceps purpurea outgroup
    ggtext::geom_richtext(
      data = tip_data %>%
        dplyr::filter(sample_type == "outgroup"),
      aes(x = label_x, y = y, label = display_label),
      inherit.aes = FALSE,
      hjust = 0,
      vjust = 0.5,
      size = 3.8,
      fill = NA,
      label.color = NA,
      label.padding = grid::unit(c(0, 0, 0, 0), "pt"),
      colour = "grey45"
    ) +

    # UFbootstrap values at a fixed distance from each node
    geom_text(
      data = node_data,
      aes(x = bootstrap_x, y = bootstrap_y, label = ufboot),
      inherit.aes = FALSE,
      colour = "black",
      size = 3.0,
      hjust = 0.5,
      vjust = 0.5
    ) +

    # Panel title
    ggtext::geom_richtext(
      data = data.frame(
        x = 0,
        y = max_y + 1.5,
        label = format_panel_title(gene_name)
      ),
      aes(x = x, y = y, label = label),
      inherit.aes = FALSE,
      hjust = 0,
      vjust = 0.5,
      size = 5.6,
      fill = NA,
      label.color = NA,
      label.padding = grid::unit(c(0, 0, 0, 0), "pt"),
      colour = "black"
    ) +

    coord_cartesian(
      xlim = c(0, label_x + max_tree_x * 0.70),
      ylim = c(0.5, max_y + 2.2),
      clip = "off"
    ) +

    ggtree::theme_tree() +

    theme(
      plot.margin = margin(t = 10, r = 15, b = 10, l = 10)
    )


  # ----------------------------------------------------------
  # DISPLAY TREE
  # ----------------------------------------------------------

  print(final_tree)


  # ----------------------------------------------------------
  # EXPORT VECTOR FILES
  # ----------------------------------------------------------

  ggsave(
    filename = file.path(
      output_dir,
      paste0(gene_name, "_individual_final.pdf")
    ),
    plot = final_tree,
    device = grDevices::pdf,
    width = 18,
    height = 24,
    units = "cm",
    bg = "white"
  )

  ggsave(
    filename = file.path(
      output_dir,
      paste0(gene_name, "_individual_final.svg")
    ),
    plot = final_tree,
    device = "svg",
    width = 18,
    height = 24,
    units = "cm",
    bg = "white"
  )

  invisible(final_tree)
}


# ------------------------------------------------------------
# 8. GENERATE ALL FIVE TREES
# ------------------------------------------------------------

final_trees <- mapply(
  FUN = process_tree_final,
  gene_name = genes,
  index = seq_along(genes),
  SIMPLIFY = FALSE
)

names(final_trees) <- genes

```


\newpage

# 3. MSC trees (ASTRAL)

## 3.1 Extract PPs and Qs support values

```{r Extract Qs & PP tables for the MSC tree}

# Set output directory
output_dir <- "../data/outputs_script_2/6_MSCtree/"

# Helper function to parse a node label that contains quartet values (q1, q2, q3)
# and returns a numeric vector c(q1, q2, q3) or c(NA, NA, NA) if it cannot parse.
parse_astral_quartets <- function(lbl) {

  if (is.null(lbl) || is.na(lbl)) return(c(NA_real_, NA_real_, NA_real_))
  lbl <- as.character(lbl)

  # If there are explicit keys like q1=..., q2=..., q3=...
  has_keys <- grepl("q1\\s*=", lbl, ignore.case = TRUE) &&
              grepl("q2\\s*=", lbl, ignore.case = TRUE) &&
              grepl("q3\\s*=", lbl, ignore.case = TRUE)

  if (has_keys) {
    q1 <- sub(".*q1\\s*=\\s*([0-9.]+).*", "\\1", lbl, ignore.case = TRUE)
    q2 <- sub(".*q2\\s*=\\s*([0-9.]+).*", "\\1", lbl, ignore.case = TRUE)
    q3 <- sub(".*q3\\s*=\\s*([0-9.]+).*", "\\1", lbl, ignore.case = TRUE)
    return(as.numeric(c(q1, q2, q3)))
  }

  # If there are three values separated by / , ; | :
  # e.g. "0.2/0.5/0.3" or "0.2,0.5,0.3"
  parts <- unlist(strsplit(lbl, "[/\\|,;:]+"))
  parts <- parts[nchar(parts) > 0]

  if (length(parts) >= 3) {
    return(suppressWarnings(as.numeric(parts[1:3])))
  }

  # If nothing works
  c(NA_real_, NA_real_, NA_real_)
}

# Extract quartet supports from an ASTRAL tree and save to Excel
extract_astral_Q_to_xlsx <- function(treefile, out_xlsx) {

  # Read ASTRAL species tree
  tr <- read.tree(treefile)

  # Node numbering compatible with ggtree
  n_tip  <- length(tr$tip.label)
  n_node <- tr$Nnode
  internal_nodes <- (n_tip + 1):(n_tip + n_node)

  # ASTRAL support is typically stored in tr$node.label
  labels <- tr$node.label

  # Parse q1/q2/q3 for each internal node label
  Qmat <- t(vapply(labels, parse_astral_quartets, numeric(3)))
  colnames(Qmat) <- c("q1", "q2", "q3")

  # Build final table
  Q_df <- tibble(node = internal_nodes,
                 q1 = Qmat[, "q1"],
                 q2 = Qmat[, "q2"],
                 q3 = Qmat[, "q3"])

  # Save to Excel
  write.xlsx(Q_df, out_xlsx, overwrite = TRUE)
  message("ASTRAL Q Excel created successfully: ", out_xlsx)

  # Helpful warning if parsing failed
  if (all(is.na(Q_df$q1)) && all(is.na(Q_df$q2)) && all(is.na(Q_df$q3))) {
    message("Warning: q1/q2/q3 are all NA. The quartet values may not be 
            stored in node labels in this treefile.")
  }
}

# Extract quartet supports from the species tree (Qs)
extract_astral_Q_to_xlsx(
  treefile = "../data/Phylogenetic_analysis/TREES/ASTRAL_SPTREE_Q",
  out_xlsx = "../data/outputs_script_2/6_MSCtree/Q_values_ASTRAL_SPTREE.xlsx")


# Helper function to parse a node label containing a posterior probability.
# Accepts formats like "0.98", "pp=0.98", "PP=0.98" and returns numeric or NA.
parse_astral_pp <- function(lbl) {

  if (is.null(lbl) || is.na(lbl)) return(NA_real_)
  lbl <- as.character(lbl)

  # If it contains pp=...
  if (grepl("pp\\s*=", lbl, ignore.case = TRUE)) {
    val <- sub(".*pp\\s*=\\s*([0-9.]+).*", "\\1", lbl, ignore.case = TRUE)
    return(suppressWarnings(as.numeric(val)))
  }

  # Otherwise, try direct numeric conversion
  suppressWarnings(as.numeric(lbl))
}


# Extract posterior probabilities from an ASTRAL tree and save to Excel. Node 
# numbering matches ggtree: internal nodes = (Ntip+1)..(Ntip+Nnode)

extract_astral_PP_to_xlsx <- function(treefile, out_xlsx) {

  # Read ASTRAL species tree
  tr <- read.tree(treefile)

  # Node numbering compatible with ggtree
  n_tip  <- length(tr$tip.label)
  n_node <- tr$Nnode
  internal_nodes <- (n_tip + 1):(n_tip + n_node)

  # PP usually stored in tr$node.label
  labels <- tr$node.label

  PP_df <- tibble(node = internal_nodes,
                  pp   = vapply(labels, parse_astral_pp, numeric(1)))

  write.xlsx(PP_df, out_xlsx, overwrite = TRUE)
  message("ASTRAL PP Excel created successfully: ", out_xlsx)

  if (all(is.na(PP_df$pp))) {
    message("Warning: pp is all NA. The PP values may not be stored in node labels 
            in this treefile.")
  }
}

# Extract posterior probability values from species tree (PP)
extract_astral_PP_to_xlsx(
  treefile = "../data/Phylogenetic_analysis/TREES/ASTRAL_SPTREE_PP",
  out_xlsx = "../data/outputs_script_2/6_MSCtree/PP_values_ASTRAL_SPTREE.xlsx")

```

## 3.2 Format, plot and save MSC tree

```{r Visualization and export of MSC tree with quartets and PP values}

# Define final labels for our samples
uz_labels <- c("UZ_11.23_8_" = "UZ_11.23(8) - Froth",
               "UZ_7.23_8_"  = "UZ_7.23(8) - Fyves",
               "UZ_55.22_2_" = "UZ_55.22(2) - Fprui",
               "UZ_9.23_4_"  = "UZ_9.23(4) - Froth",
               "UZ_55.22_4_" = "UZ_55.22(4) - Fprui",
               "UZ_44.22_1_" = "UZ_44.22(1) - Fnigr",
               "UZ_13.22_5_" = "UZ_13.22(5) - Flamb",
               "UZ_4.22_1_"  = "UZ_4.22(1) - Ftrich",
               "UZ_7.23_11_" = "UZ_7.23(11) - Fyves",
               "UZ_44.22_4_" = "UZ_44.22(4) - Fnigr")


# Clean raw tip labels
clean_tip_label <- function(x) {

  x <- trimws(x)

  is_uz <- x %in% names(uz_labels)
  x[is_uz] <- unname(uz_labels[x[is_uz]])

  x[!is_uz] <- gsub("_", " ", x[!is_uz])
  x[!is_uz] <- gsub("\\bEpichlo[eë]\\b", "E.", x[!is_uz])
  x[!is_uz] <- gsub("\\bClaviceps\\b", "C.", x[!is_uz])
  x[!is_uz] <- gsub("^E\\.\\s*", "E. ", x[!is_uz])
  x[!is_uz] <- gsub("^C\\.\\s*", "C. ", x[!is_uz])
  x[!is_uz] <- gsub("^E\\. typhinatyphina\\b", "E. typhina subsp. typhina", x[!is_uz])
  x[!is_uz] <- gsub("^E\\. typhinaclarkii\\b", "E. typhina subsp. clarkii", x[!is_uz])
  x[!is_uz] <- gsub("^E\\. typhinapoae\\b", "E. typhina subsp. poae", x[!is_uz])

  x <- gsub("\\s+", " ", x)

  trimws(x)
}


# Identify our samples and outgroup
is_our_sample <- function(x) {
  grepl("^UZ_", x)
}

is_outgroup <- function(x) {
  grepl("^C\\. purpurea\\b", x, ignore.case = TRUE)
}


# Format tip labels
format_tip_label <- function(x) {

  # Our samples: complete label in bold
  if (is_our_sample(x)) {
    return(paste0("**", x, "**"))
  }

  # E. typhina subspecies
  if (grepl("^E\\. typhina subsp\\. [a-z-]+\\b", x)) {

    subspecies <- stringr::str_extract(x, "(?<=subsp\\. )[a-z-]+")
    rest <- trimws(sub(paste0("^E\\. typhina subsp\\. ", subspecies), "", x))

    return(paste0("*E. typhina* subsp. *", subspecies, "*",
                  ifelse(rest == "", "", paste0(" ", rest))))
  }

  # Standard Epichloë names
  if (grepl("^E\\. [a-z-]+\\b", x)) {

    species <- stringr::str_extract(x, "^E\\. [a-z-]+")
    rest <- trimws(sub(paste0("^", gsub("\\.", "\\\\.", species)), "", x))

    return(paste0("*", species, "*",
                  ifelse(rest == "", "", paste0(" ", rest))))
  }

  # C. purpurea outgroup
  if (grepl("^C\\. purpurea\\b", x)) {

    rest <- trimws(sub("^C\\. purpurea", "", x))

    return(paste0("*C. purpurea*",
                  ifelse(rest == "", "", paste0(" ", rest))))
  }

  x
}


# Read ASTRAL trees and support tables
q_tree_path <- "../data/Phylogenetic_analysis/TREES/ASTRAL_SPTREE_Q"
pp_tree_path <- "../data/Phylogenetic_analysis/TREES/ASTRAL_SPTREE_PP"

q_file <- "../data/outputs_script_2/6_MSCtree/Q_values_ASTRAL_SPTREE.xlsx"
pp_file <- "../data/outputs_script_2/6_MSCtree/PP_values_ASTRAL_SPTREE.xlsx"


if (!file.exists(q_tree_path)) stop("ASTRAL Q tree not found.")
if (!file.exists(pp_tree_path)) stop("ASTRAL PP tree not found.")
if (!file.exists(q_file)) stop("ASTRAL Q support file not found.")
if (!file.exists(pp_file)) stop("ASTRAL PP support file not found.")

if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

tree_Q <- ape::read.tree(q_tree_path)
tree_PP <- ape::read.tree(pp_tree_path)


# Check that both trees have the same dimensions
stopifnot(ape::Ntip(tree_Q) == ape::Ntip(tree_PP),
          tree_Q$Nnode == tree_PP$Nnode)


# Clean tip labels
tree_Q$tip.label <- vapply(tree_Q$tip.label, clean_tip_label, character(1))


# Create base tree
base_astral <- ggtree::ggtree(tree_Q, layout = "rectangular", linewidth = 0.50)

tree_data <- base_astral$data

max_tree_x <- max(tree_data$x, na.rm = TRUE)
max_y <- max(tree_data$y, na.rm = TRUE)


# Prepare tip labels
tip_data <- tree_data %>%
  dplyr::filter(isTip) %>%
  dplyr::mutate(display_label = vapply(label, format_tip_label, character(1)),
                sample_type = dplyr::case_when(
                  vapply(label, is_our_sample, logical(1)) ~ "sample",
                  vapply(label, is_outgroup, logical(1)) ~ "outgroup",
                  TRUE ~ "reference"))


# Prepare quartet and PP support values
q_values <- openxlsx::read.xlsx(q_file) %>%
  dplyr::mutate(node = as.numeric(node), q1 = as.numeric(q1),
                q2 = as.numeric(q2), q3 = as.numeric(q3))

pp_values <- openxlsx::read.xlsx(pp_file) %>%
  dplyr::mutate(node = as.numeric(node), pp = as.numeric(pp))

support_data <- tree_data %>%
  dplyr::filter(!isTip) %>%
  dplyr::left_join(q_values, by = "node") %>%
  dplyr::left_join(pp_values, by = "node") %>%
  dplyr::filter(!is.na(q1) | !is.na(q2) | !is.na(q3) | !is.na(pp))


# Assign TIFF node numbers
tiff_node_numbers <- tibble::tribble(
  ~node, ~display_node,
  38,  1,
  39,  2,
  40,  3,
  41,  4,
  42,  5,
  59,  6,
  60,  7,
  61,  8,
  62,  9,
  63, 10,
  64, 11,
  65, 12,
  66, 13,
  67, 14,
  68, 15,
  69, 16,
  43, 17,
  44, 18,
  45, 19,
  46, 20,
  47, 21,
  48, 22,
  49, 23,
  50, 24,
  51, 25,
  52, 26,
  53, 27,
  54, 28,
  55, 29,
  56, 30,
  57, 31,
  58, 32)

support_data <- support_data %>%
  dplyr::left_join(tiff_node_numbers, by = "node")


# Format PP values
support_data <- support_data %>%
  dplyr::mutate(pp_label = dplyr::case_when(is.na(pp) ~ "",
                                            TRUE ~ sprintf("%.2f", pp)))


# Define support-symbol positions
# Bring tip labels closer to the tree
label_x <- max_tree_x + max_tree_x * 0.03

# Larger circular pies
pie_radius <- 0.45

# Default PP position
default_pp_dx <- -max_tree_x * 0.025
default_pp_dy <- 0.35

# Node numbers: fixed position to the left of each node
node_number_dx <- max_tree_x * 0.042
node_number_dy <- 0.00


# Define manual PP-label adjustments
# dx: negative = left and positive = right
# dy: positive = up and negative = down
# Use NA_real_ to keep the default value for that coordinate.

pp_adjustments <- tibble::tibble(node = numeric(), dx = numeric(), dy = numeric())

pp_adjustments <- tibble::tibble(
  display_node = 1:32, dx = -0.55,
  dy = dplyr::if_else(display_node %in% c(9, 19), -0.67, 0.67))


# Calculate final PP and node-number positions
support_data <- support_data %>%
  dplyr::select(-dplyr::any_of(c("dx", "dy", "dx.x", "dx.y", "dy.x", "dy.y",
                                 "pp_x", "pp_y", "number_x", "number_y"))) %>%
  dplyr::mutate(pp_x = x + default_pp_dx, pp_y = y + default_pp_dy,
                number_x = x + node_number_dx, number_y = y + node_number_dy) %>%
  dplyr::left_join(pp_adjustments, by = "display_node") %>%
  dplyr::mutate(pp_x = dplyr::if_else(!is.na(dx), x + dx, pp_x),
                pp_y = dplyr::if_else(!is.na(dy), y + dy, pp_y))


# Optional diagnostic table
pp_node_table <- support_data %>%
  dplyr::select(node, display_node, pp_label, x, y) %>%
  dplyr::arrange(dplyr::desc(y))


# Define section annotations
# Retrieve y position of a tip from its final cleaned label
get_tip_y <- function(label_name) {

  result <- tip_data %>%
    dplyr::filter(label == label_name) %>%
    dplyr::pull(y)

  if (length(result) != 1) stop("Tip label not uniquely found: ", label_name)

  result
}


# Bars on the right side of the tree
section_data <- tibble::tribble(
  ~section, ~top_tip, ~bottom_tip,
  "*Festuca*", "UZ_7.23(11) - Fyves", "UZ_7.23(8) - Fyves",
  "*Aulaxyper*", "UZ_9.23(4) - Froth", "UZ_11.23(8) - Froth",
  "*Aulaxyper*<br>(gr. *rubra*)", "UZ_44.22(4) - Fnigr", "UZ_55.22(2) - Fprui")

section_data <- section_data %>%
  dplyr::rowwise() %>%
  dplyr::mutate(y_top = get_tip_y(top_tip) + 0.40,
                y_bottom = get_tip_y(bottom_tip) - 0.40,
                y_mid = mean(c(y_top, y_bottom))) %>%
  dplyr::ungroup()

section_bar_x <- label_x + max_tree_x * 0.45
section_text_x <- section_bar_x + max_tree_x * 0.05


# Define pie colours
quartet_colours <- c("q1" = "#855885", "q2" = "#F1A340", "q3" = "#F7F7F7")


# Construct final ASTRAL tree
ASTRAL_tree_final <- base_astral +
  
  scatterpie::geom_scatterpie(data = support_data, aes(x = x, y = y, r = pie_radius),
                              cols = c("q1", "q2", "q3"), colour = "black",
                              linewidth = 0.35, inherit.aes = FALSE) +
  scale_fill_manual(values = c("q1" = "#855885", "q2" = "#F1A340", "q3" = "#F7F7F7"),
                    breaks = c("q1", "q2", "q3")) +
  geom_text(data = support_data, aes(x = pp_x, y = pp_y, label = pp_label),
            inherit.aes = FALSE, colour = "black", size = 3.0,
            hjust = 0.5, vjust = 0.5) +
  geom_text(data = support_data, aes(x = number_x, y = number_y, label = display_node),
            inherit.aes = FALSE, colour = "grey16", size = 2.8,
            hjust = 0, vjust = 0.5) +
  ggtext::geom_richtext(data = tip_data %>% dplyr::filter(sample_type == "reference"),
                        aes(x = label_x, y = y, label = display_label),
                        inherit.aes = FALSE, hjust = 0, vjust = 0.5, size = 3.9,
                        fill = NA, label.color = NA, colour = "black",
                        label.padding = grid::unit(c(0, 0, 0, 0), "pt")) +
  ggtext::geom_richtext(data = tip_data %>% dplyr::filter(sample_type == "sample"),
                        aes(x = label_x, y = y, label = display_label),
                        inherit.aes = FALSE, hjust = 0, vjust = 0.5, size = 3.9,
                        fill = NA, label.color = NA, colour = "black",
                        label.padding = grid::unit(c(0, 0, 0, 0), "pt")) +
  ggtext::geom_richtext(data = tip_data %>% dplyr::filter(sample_type == "outgroup"),
                        aes(x = label_x, y = y, label = display_label),
                        inherit.aes = FALSE, hjust = 0, vjust = 0.5, size = 3.9,
                        fill = NA, label.color = NA, colour = "grey50",
                        label.padding = grid::unit(c(0, 0, 0, 0), "pt")) +
  geom_segment(data = section_data,
               aes(x = section_bar_x, xend = section_bar_x, y = y_bottom, 
                   yend = y_top), inherit.aes = FALSE, colour = "black", 
               linewidth = 1.2) +
  ggtext::geom_richtext(data = section_data,
                        aes(x = section_text_x, y = y_mid, label = section),
                        inherit.aes = FALSE, hjust = 0, vjust = 0.5, size = 4.2,
                        fill = NA, label.color = NA, colour = "black",
                        label.padding = grid::unit(c(0, 0, 0, 0), "pt")) +
  coord_cartesian(xlim = c(min(tree_data$x, na.rm = TRUE) - max_tree_x * 0.02,
                           section_text_x + max_tree_x * 0.12),
                  ylim = c(0.3, max_y + 0.4), clip = "off") +
  ggtree::theme_tree() +
  theme(legend.position = "none",
        plot.margin = margin(t = 6, r = 6, b = 6, l = 6))

# Display final tree
print(ASTRAL_tree_final)


# Export final tree
output_dir <- "../data/outputs_script_2/6_MSCtree/"

ggsave(filename = file.path(output_dir, "ASTRAL.pdf"), plot = ASTRAL_tree_final,
       device = grDevices::pdf, width = 19, height = 21, units = "cm", bg = "white")

ggsave(filename = file.path(output_dir, "ASTRAL.svg"), plot = ASTRAL_tree_final,
       device = "svg", width = 18.3, height = 23.8, units = "cm", bg = "white")

```

\newpage


# 4. ML concatenated tree (IQTREE2)

## 4.1 Extract UFbootstrap and scFL support values

```{r Extract SCFL and UFbootstrap values for the ML tree}

# Extract node, sCF (scfl) and Label (ufboot) directly from IQ-TREE cf.stat file
# and save the values into an excel file
extract_scfl_and_ufboot_from_cfstat <- function(cf_stat_file, out_xlsx) {

  # Read cf.stat ignoring comment lines starting with '#'
  stat <- read.delim(cf_stat_file, sep = "\t", header = TRUE, comment.char = "#",
                     stringsAsFactors = FALSE, check.names = FALSE)

  # Validate required columns
  required_cols <- c("ID", "sCF", "Label")
  if (!all(required_cols %in% names(stat))) {
    stop("cf.stat must contain columns: ID, sCF, Label")
  }

  # Select and rename columns to match your plotting pipeline
  out <- stat %>% dplyr::transmute(node = as.numeric(ID) +1,
                                   scfl = as.numeric(sCF),
                                   ufboot = as.numeric(Label))

  # Save Excel
  write.xlsx(out, out_xlsx, overwrite = TRUE)

  message("Excel with node, scfl and ufboot created: ", out_xlsx)
}

# Extract the values and save them into an excel file
extract_scfl_and_ufboot_from_cfstat(
  cf_stat_file = "../data/Phylogenetic_analysis/TREES/ML_TREE.cf.stat",
  out_xlsx = "../data/outputs_script_2/7_MLtree/UFBoot_and_scfl_values_MLtree.xlsx")

```


## 4.2 Format, plot and save ML tree

```{r Visualization and export of ML concatenated tree with UFboot and SCFL values}

# Define final labels for our samples
uz_labels <- c("UZ_11.23_8_" = "UZ_11.23(8) - Froth",
               "UZ_7.23_8_"  = "UZ_7.23(8) - Fyves",
               "UZ_55.22_2_" = "UZ_55.22(2) - Fprui",
               "UZ_9.23_4_"  = "UZ_9.23(4) - Froth",
               "UZ_55.22_4_" = "UZ_55.22(4) - Fprui",
               "UZ_44.22_1_" = "UZ_44.22(1) - Fnigr",
               "UZ_13.22_5_" = "UZ_13.22(5) - Flamb",
               "UZ_4.22_1_"  = "UZ_4.22(1) - Ftrich",
               "UZ_7.23_11_" = "UZ_7.23(11) - Fyves",
               "UZ_44.22_4_" = "UZ_44.22(4) - Fnigr")


# Clean raw tip labels
clean_tip_label <- function(x) {

  x <- trimws(x)

  is_uz <- x %in% names(uz_labels)
  x[is_uz] <- unname(uz_labels[x[is_uz]])

  x[!is_uz] <- gsub("_", " ", x[!is_uz])
  x[!is_uz] <- gsub("\\bEpichlo[eë]\\b", "E.", x[!is_uz])
  x[!is_uz] <- gsub("\\bClaviceps\\b", "C.", x[!is_uz])
  x[!is_uz] <- gsub("^E\\.\\s*", "E. ", x[!is_uz])
  x[!is_uz] <- gsub("^C\\.\\s*", "C. ", x[!is_uz])
  x[!is_uz] <- gsub("^E\\. typhinatyphina\\b",
                    "E. typhina subsp. typhina", x[!is_uz])
  x[!is_uz] <- gsub("^E\\. typhinaclarkii\\b",
                    "E. typhina subsp. clarkii", x[!is_uz])
  x[!is_uz] <- gsub("^E\\. typhinapoae\\b",
                    "E. typhina subsp. poae", x[!is_uz])

  x <- gsub("\\s+", " ", x)

  trimws(x)
}


# Identify our samples and outgroup
is_our_sample <- function(x) {
  grepl("^UZ_", x)
}

is_outgroup <- function(x) {
  grepl("^C\\. purpurea\\b", x, ignore.case = TRUE)
}


# Format tip labels
format_tip_label <- function(x) {

  if (is_our_sample(x)) {
    return(paste0("**", x, "**"))
  }

  if (grepl("^E\\. typhina subsp\\. [a-z-]+\\b", x)) {

    subspecies <- stringr::str_extract(x, "(?<=subsp\\. )[a-z-]+")
    rest <- trimws(sub(paste0("^E\\. typhina subsp\\. ", subspecies), "", x))

    return(paste0("*E. typhina* subsp. *", subspecies, "*",
                  ifelse(rest == "", "", paste0(" ", rest))))
  }

  if (grepl("^E\\. [a-z-]+\\b", x)) {

    species <- stringr::str_extract(x, "^E\\. [a-z-]+")
    rest <- trimws(sub(paste0("^", gsub("\\.", "\\\\.", species)), "", x))

    return(paste0("*", species, "*",
                  ifelse(rest == "", "", paste0(" ", rest))))
  }

  if (grepl("^C\\. purpurea\\b", x)) {

    rest <- trimws(sub("^C\\. purpurea", "", x))

    return(paste0("*C. purpurea*",
                  ifelse(rest == "", "", paste0(" ", rest))))
  }

  x
}


# Read and prepare ML tree
treefile_path <- "../data/Phylogenetic_analysis/TREES/ML_TREE.treefile"
supportfile_path <- "../data/outputs_script_2/7_MLtree/UFBoot_and_scfl_values_MLtree.xlsx"

if (!file.exists(treefile_path)) stop("Tree file not found: ", treefile_path)
if (!file.exists(supportfile_path)) stop("Support file not found: ", supportfile_path)
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

treeML <- ape::read.tree(treefile_path)
treeML$edge.length <- sqrt(sqrt(treeML$edge.length))
treeML$tip.label <- vapply(treeML$tip.label, clean_tip_label, character(1))


# Create base tree
base_treeML <- ggtree::ggtree(treeML, layout = "rectangular", linewidth = 0.50)
tree_data <- base_treeML$data


# Prepare tip labels
tip_data <- tree_data %>%
  dplyr::filter(isTip) %>%
  dplyr::mutate(display_label = vapply(label, format_tip_label, character(1)),
                sample_type = dplyr::case_when(
                  vapply(label, is_our_sample, logical(1)) ~ "sample",
                  vapply(label, is_outgroup, logical(1)) ~ "outgroup",
                  TRUE ~ "reference"))


# Prepare node support values
bootscfl <- openxlsx::read.xlsx(supportfile_path)

bootscfl <- bootscfl %>%
  dplyr::mutate(node = as.numeric(node), ufboot = as.numeric(ufboot),
                scfl = as.numeric(scfl))

node_data <- tree_data %>%
  dplyr::filter(!isTip) %>%
  dplyr::left_join(bootscfl, by = "node") %>%
  dplyr::filter(!is.na(ufboot), !is.na(scfl)) %>%
  dplyr::mutate(support_label = paste0(sprintf("%.0f", ufboot), "/",
                                       sprintf("%.1f", scfl)))


# Define general plot coordinates
max_tree_x <- max(tree_data$x, na.rm = TRUE)
max_y <- max(tree_data$y, na.rm = TRUE)

label_x <- max_tree_x + max_tree_x * 0.20
label_line_end <- label_x - max_tree_x * 0.015


# Define default support-label position
# Default position for all support values
default_support_dx <- -max_tree_x * 0.030
default_support_dy <- 0.28
default_support_hjust <- 1


# Define manual support-label adjustments
# dx: negative = move left and positive = move right
# dy: positive = move up and negative = move down
support_adjustments <- tibble::tibble(node = numeric(), dx = numeric(),
                                      dy = numeric())

support_adjustments <- tibble::tribble(
  ~node, ~dx, ~dy, 
  37, -0.05, 0.45, 
  38, -0.05, 0.45, 
  39, -0.05, 0.45, 
  40, -0.05, 0.45, 
  41, -0.05, 0.45, 
  45, -0.05, 0.45, 
  47, -0.05, 0.45, 
  48, -0.05, 0.45, 
  49, -0.05, 0.45,
  50, -0.05, 0.45, 
  51, -0.05, 0.45, 
  52, -0.05, 0.45, 
  53, -0.05, 0.45, 
  67, 0.6, -0.02, 
  68, 0.6, -0.02, 
  57, 0.55, -0.02, 
  66, 0.6, -0.02, 
  65, 0.6, -0.02, 
  58, 0.55, -0.55, 
  59, -0.05, 0.45, 
  64, 0.6, -0.02, 
  60, -0.05, 0.45, 
  61, -0.05, 0.45, 
  62, -0.05, 0.45, 
  63, -0.05, 0.45, 
  42, 0.55, -0.02, 
  43, 0.6, -0.02, 
  44, 0.6, -0.02, 
  46, 0.55, -0.02, 
  56, 0.55, -0.02, 
  55, 0.55, -0.02, 
  54, 0.55, -0.02)


# Calculate final support-label positions
# Remove previous manual-adjustment columns if this block is rerun
node_data <- node_data %>%
  dplyr::select(-dplyr::any_of(c("dx", "dy", "dx.x", "dx.y",
                                 "dy.x", "dy.y")))

# Apply default positions and manual corrections
node_data <- node_data %>%
  dplyr::mutate(support_x = x + default_support_dx,
                support_y = y + default_support_dy) %>%
  dplyr::left_join(support_adjustments, by = "node") %>%
  dplyr::mutate(support_x = dplyr::if_else(!is.na(dx), x + dx, support_x),
                support_y = dplyr::if_else(!is.na(dy), y + dy, support_y))


# Inspect node numbers
# Run this table when you need to identify which support label
# corresponds to each internal node.
support_node_table <- node_data %>%
  dplyr::select(node, support_label, x, y) %>%
  dplyr::arrange(dplyr::desc(y))

print(support_node_table)


# Debug tree with node numbers
ML_tree_debug <- base_treeML +

  geom_text(data = node_data, aes(x = x, y = y, label = node),
            inherit.aes = FALSE, colour = "red", size = 3,
            nudge_x = max_tree_x * 0.015) +

  coord_cartesian(xlim = c(0, label_x + max_tree_x * 0.85),
                  ylim = c(0.5, max_y + 0.8), clip = "off") +

  ggtree::theme_tree()

# Uncomment when needed:
print(ML_tree_debug)


# Construct final ML tree
ML_tree_final <- base_treeML +

  # Black dotted lines connecting tips and labels
  geom_segment(data = tip_data,
               aes(x = x, xend = label_line_end, y = y, yend = y),
               inherit.aes = FALSE, colour = "black", linewidth = 0.40,
               linetype = "dotted") +

  # Reference taxa
  ggtext::geom_richtext(data = tip_data %>%
                          dplyr::filter(sample_type == "reference"),
                        aes(x = label_x, y = y, label = display_label),
                        inherit.aes = FALSE, hjust = 0, vjust = 0.5, size = 3.8,
                        fill = NA, label.color = NA, colour = "black",
                        label.padding = grid::unit(c(0, 0, 0, 0), "pt")) +

  # Our samples
  ggtext::geom_richtext(data = tip_data %>%
                          dplyr::filter(sample_type == "sample"),
                        aes(x = label_x, y = y, label = display_label),
                        inherit.aes = FALSE, hjust = 0, vjust = 0.5, size = 3.8,
                        fill = NA, label.color = NA, colour = "black",
                        label.padding = grid::unit(c(0, 0, 0, 0), "pt")) +

  # Claviceps purpurea outgroup
  ggtext::geom_richtext(data = tip_data %>%
                          dplyr::filter(sample_type == "outgroup"),
                        aes(x = label_x, y = y, label = display_label),
                        inherit.aes = FALSE, hjust = 0, vjust = 0.5, size = 3.8,
                        fill = NA, label.color = NA, colour = "grey45",
                        label.padding = grid::unit(c(0, 0, 0, 0), "pt")) +

  # UFbootstrap / sCFL support values
  geom_text(data = node_data,
            aes(x = support_x, y = support_y, label = support_label,
                hjust = default_support_hjust),
            inherit.aes = FALSE, colour = "black", size = 3.0, vjust = 0.5) +

  coord_cartesian(xlim = c(0, label_x + max_tree_x * 0.85),
                  ylim = c(0.5, max_y + 0.8), clip = "off") +

  ggtree::theme_tree() +

  theme(plot.margin = margin(t = 10, r = 15, b = 10, l = 10))


# Display final tree
print(ML_tree_final)


# Export final tree
output_dir <- "../data/outputs_script_2/7_MLtree/"

ggsave(filename = file.path(output_dir, "MLtree.pdf"), plot = ML_tree_final,
       device = grDevices::pdf, width = 18, height = 24, units = "cm",
       bg = "white")

ggsave(filename = file.path(output_dir, "MLtree.svg"), plot = ML_tree_final,
       device = "svg", width = 18, height = 24, units = "cm", bg = "white")

```


```{r End of script}

# Notify script completion and indicate the location of the output files
message(
  "\n\nScript executed successfully.\n\n",
  "All output files have been saved to the corresponding output directory",
  "within the downloaded GitHub repository.\n\n"
)

```
